---
title: "SQ2: Does Neighborhood Income Affect 311 Resolution Times?"
author: "Raúl J. Solá Navarro"
date: today
format:
  html:
    toc: true
    number-sections: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    self-contained: true
    output-file: "individual_report_raul.html"
output-dir: docs
---

## Overarching Question

What factors drive 311 service request response and resolution times 
and do they suggest segmented prioritization of issues?

## Specific Question

Do lower-income NYC census tracts experience longer 311 service request 
resolution times, and does this pattern hold across complaint types?

This report examines whether neighborhood income, measured as median 
household income at the census tract level, is associated with how 
quickly NYC agencies resolve 311 service requests. Using 12.6 million 
closed requests from 2022-2025 and ACS income data, we find a modest 
but meaningful raw disparity: the lowest-income neighborhoods wait 
about 43% longer than the wealthiest. However, after controlling for 
complaint type and borough, much of that gap is explained by *what* 
neighborhoods request, not *where* they are.

## Data Sources

### NYC 311 Service Requests (2022–2025)
Pulled via the NYC Open Data Socrata API. Filtered to closed requests 
with valid coordinates and resolution times between 0 and 8,760 hours 
(1 year). Final cleaned dataset: 12.6 million requests.

### American Community Survey (tidycensus)
Median household income (variable B19013_001) at the census tract level 
for all five NYC boroughs, from the 2019-2023 5-year ACS. Retrieved via 
the `tidycensus` R package with tract geometries for spatial joining.
2,327 census tracts; 5.8% excluded due to missing income estimates.

## Data Acquisition

All data is loaded from pre-processed local files. Raw data was acquired 
via the NYC Open Data Socrata API (311 requests) and the `tidycensus` R 
package (ACS income estimates). See `data_pull.R` and 
`data_clean_census.R` for the full acquisition code.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(sf)
library(tidycensus)
library(kableExtra)
library(patchwork)

# Load pre-computed results
df               <- readRDS("data/clean_311_final.rds")
nyc_income       <- readRDS("data/nyc_income_tracts.rds")
boot_ci          <- readRDS("data/boot_ci.rds")
model_results    <- readRDS("data/model_results.rds")
tract_summary    <- readRDS("data/tract_summary.rds")
quintile_summary <- readRDS("data/quintile_summary.rds")
```

## Data Quality

The 311 dataset required removal of 466,215 records (3.6%) with missing 
dates, negative resolution times, or unspecified boroughs. Of the 
2,327 census tracts, 5.8% had missing income estimates and 11% had 
high margins of error (CV > 50%). The quintile-based analysis approach 
is robust to individual tract-level noise, since bucketing tracts into 
five groups is far less sensitive to noisy individual estimates than 
treating income as a precise continuous measure.

## Exploratory Data Analysis

Resolution times are heavily right-skewed having most requests resolve 
within hours, but a long tail of slow resolutions pulls the mean to 
236 hours. The table and charts below summarize the key patterns.

```{r}
#| message: false
#| warning: false

quintile_summary |>
  mutate(
    income_quintile = c("Q1 (Lowest)", "Q2", "Q3", "Q4", "Q5 (Highest)"),
    median_hours = round(median_hours, 2),
    mean_hours = round(mean_hours, 1),
    n = scales::comma(n)
  ) |>
  kable(
    caption = "Median and mean 311 resolution time by neighborhood income quintile",
    col.names = c("Income Quintile", "Median Hours",
                  "Mean Hours", "Requests"),
    align = "lrrr"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = TRUE) |>
  column_spec(1, width = "25%") |>
  column_spec(2, width = "25%") |>
  column_spec(3, width = "25%") |>
  column_spec(4, width = "25%")
```

```{r}
#| message: false
#| warning: false
#| fig-cap: "Resolution time is heavily right-skewed; log scale reveals a cleaner distribution"

df |>
  slice_sample(n = 500000) |>
  ggplot(aes(x = resolution_hours)) +
  geom_histogram(bins = 100, fill = "steelblue", color = NA) +
  scale_x_log10(labels = scales::comma) +
  labs(
    title = "Distribution of 311 Resolution Times",
    x = "Resolution Time (hours, log scale)",
    y = "Count"
  ) +
  theme_minimal()
```

```{r}
#| message: false
#| warning: false
#| fig-cap: "Lower-income quintiles generally show longer median resolution times"

quintile_summary |>
  mutate(income_quintile = factor(income_quintile,
    labels = c("Q1\n(Lowest)", "Q2", "Q3", "Q4", "Q5\n(Highest)"))) |>
  ggplot(aes(x = income_quintile, y = median_hours,
             fill = income_quintile)) +
  geom_col() +
  scale_fill_manual(values = c(
    "Q1\n(Lowest)" = "#084594",
    "Q2"           = "#2171b5",
    "Q3"           = "#4292c6",
    "Q4"           = "#6baed6",
    "Q5\n(Highest)"= "#c6dbef"
  )) +
  labs(
    title = "Median 311 Resolution Time by Neighborhood Income Quintile",
    x = "Income Quintile",
    y = "Median Resolution Time (hours)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
#| message: false
#| warning: false
#| fig-cap: "Each point is one census tract (min. 50 requests). The loess curve shows a non-linear relationship between income and resolution time."

tract_summary_income <- df |>
  filter(!is.na(geoid), !is.na(median_income)) |>
  group_by(geoid, median_income) |>
  summarise(median_hours = median(resolution_hours),
            n = n(), .groups = "drop") |>
  filter(n >= 50)

ggplot(tract_summary_income,
       aes(x = median_income, y = median_hours)) +
  geom_point(alpha = 0.3, size = 0.8, color = "steelblue") +
  geom_smooth(method = "loess", color = "red", se = TRUE) +
  scale_x_continuous(labels = scales::dollar) +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "Tract-Level Income vs. Median 311 Resolution Time",
    x = "Median Household Income",
    y = "Median Resolution Time (hours, log scale)",
    caption = "Loess smoothing line with 95% CI"
  ) +
  theme_minimal()
```

## Analysis

### Regression Model

To isolate the income effect, we fit an OLS regression of 
log(resolution time) on income quintile, controlling for complaint 
type and borough. Logging the outcome reduces the influence of extreme 
outliers and makes coefficients interpretable as approximate percentage 
differences.

```{r}
#| message: false
#| warning: false

model_results |>
  mutate(
    term = c("Q2", "Q3", "Q4", "Q5 (Highest)"),
    estimate = round(estimate, 4),
    pct_difference = paste0(pct_difference, "%"),
    p.value = ifelse(p.value < 0.001, "< 0.001", formatC(p.value, format = "e", digits = 2))
  ) |>
  select(term, estimate, pct_difference, p.value) |>
  kable(
    caption = "Income quintile coefficients from OLS regression, relative to Q1",
    col.names = c("Quintile vs Q1", "Estimate (log hrs)",
                  "% Difference", "P-value"),
    align = "lrrr"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = TRUE) |>
  column_spec(1, width = "25%") |>
  column_spec(2, width = "25%") |>
  column_spec(3, width = "25%") |>
  column_spec(4, width = "25%")
```

### Bootstrap Confidence Intervals

To assess uncertainty without relying on OLS standard errors, we 
bootstrap the regression using 200 iterations on a 10,000-row sample. 
The resulting 95% confidence intervals confirm that the income effect, 
while statistically detectable, is modest in magnitude.

```{r}
#| message: false
#| warning: false
#| fig-cap: "Bootstrap 95% CIs for income quintile effects relative to Q1"

boot_ci |>
  mutate(term = recode(term,
    "income_quintile2" = "Q2",
    "income_quintile3" = "Q3",
    "income_quintile4" = "Q4",
    "income_quintile5" = "Q5 (Highest)"
  )) |>
  ggplot(aes(x = term, y = mean_estimate,
             ymin = ci_lower, ymax = ci_upper)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbar(width = 0.2, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  labs(
    title = "Effect of Income Quintile on Log Resolution Time",
    subtitle = "Relative to Q1 (Lowest Income), controlling for complaint type and borough",
    x = "Income Quintile",
    y = "Estimated Effect (log hours)",
    caption = "Bootstrap 95% CI, 200 iterations, n=10,000 sample"
  ) +
  theme_minimal()
```

## Maps

The two maps below allow direct visual comparison of neighborhood 
income and 311 resolution times across NYC census tracts. If income 
were a strong driver of resolution time, the spatial patterns would 
closely mirror each other.

```{r}
#| message: false
#| warning: false
#| fig-cap: "Median household income (left) and 311 resolution time (right) by census tract"
#| fig-width: 12

tract_map <- nyc_income |>
  left_join(tract_summary, by = "geoid") |>
  st_transform(crs = 4326)

map_income <- ggplot(tract_map) +
  geom_sf(aes(fill = estimate), color = NA) +
  scale_fill_viridis_c(
    option = "viridis",
    labels = scales::dollar,
    name = "Median Income"
  ) +
  labs(title = "Median Household Income",
       subtitle = "NYC, 2019–2023 ACS") +
  theme_void()

map_resolution <- ggplot(tract_map) +
  geom_sf(aes(fill = median_hours), color = NA) +
  scale_fill_viridis_c(
    option = "magma",
    direction = -1,
    trans = "log10",
    labels = scales::comma,
    name = "Median Hours\n(log scale)"
  ) +
  labs(title = "Median 311 Resolution Time",
       subtitle = "NYC, 2022–2025",
       caption = "Tracts with fewer than 50 requests excluded") +
  theme_void()

map_income + map_resolution
```

## Findings

The raw data shows that Q1 (lowest income) tracts have a median 
resolution time of 8.64 hours versus 6.06 hours for Q5 (highest 
income), resulting in a 43% gap. However, after controlling for 
complaint type and borough, the income effect weakens considerably 
and becomes non-linear. This suggests that much of the apparent 
disparity is driven by the *types* of complaints filed in different 
neighborhoods rather than income itself.

Notable exceptions include HEAT/HOT WATER complaints, which show 
little income variation due to strict legal [SLAs](https://data.cityofnewyork.us/City-Government/311-Service-Level-Agreements/cs9t-e3x8/about_data), 
and UNSANITARY CONDITION complaints, where wealthy tracts show 
dramatically longer resolution times. The spatial maps reinforce the 
finding that resolution time patterns do not closely mirror the income 
map, suggesting other factors are at play.

## Limitations

- ACS income estimates reflect 2019-2023 averages and may not 
  capture recent changes
- "Closed" status may reflect administrative closure rather than 
  genuine resolution
- Bootstrap inference used a 10,000-row sample due to computational 
  constraints
- 11% of tracts have high margins of error (CV > 50%) in income estimates