---
title: "STA 9750 Mini-Project #02: How Do You Do 'You Do You'?"
author: "Raúl J. Solá Navarro"
date: "2026-04-03"
date-modified: "2026-10-06"
language:
  title-block-published: "Originally Published"
  title-block-modified: "Republished"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-fold: true
    code-summary: "View R Code"
    embed-resources: true
    theme: flatly
    mainfont: "Georgia"
    fontsize: 1.05em
    linestretch: 1.6
---

## Introduction

The American Time Use Survey (ATUS), a joint initiative of the US Bureau 
of Labor Statistics and the US Census Bureau, provides a look into 
how Americans allocate their time. In this report, we analyze over two decades 
of ATUS microdata (2003–2024) looking for patterns regarding how different 
demographic groups spend their days. We then apply these findings to design 
time-use profiles for a new time-tracking app targeting specific user 
demographics.

::: {.callout-note title="Republication note (October 6, 2026)" collapse="true"}
This report was first published on April 3, 2026. It was republished on 
October 6, 2026 after an audit of the analysis code found errors that changed 
several results. Every table, chart, and number below was regenerated. The 
main corrections:

- **Employment status** now follows the BLS definitions of the ATUS labor 
  force variable. Previously, "not in labor force" was labeled "retired", and 
  unemployed respondents were labeled "not employed" in the stay-at-home dad 
  and career-changer groups.
- **Child care** now uses the ATUS categories for caring for household 
  children. Previously, a text filter matched "Personal Care" (mostly sleep) 
  instead.
- **Averages are per respondent per day**, including respondents who did not 
  do the activity. Several tables previously averaged the length of each 
  logged activity instead.
- **Respondent demographics** are now linked using the ATUS person line number 
  and validated against the ATUS roster.
- **Profile targets** now have a stated direction (a minimum or a limit) set 
  near what peers actually do, and the table reports the share of the group 
  that meets each target. The stay-at-home dad group is defined more tightly.
- **Smaller fixes:** transit location codes, the diary month used for the 
  summer analysis, college enrollment, the sports count, and combined 
  multi-year activity codes that were previously dropped.
:::

## Data Acquisition

<!-- Task 1, 2 and 3: Data Acquisition, Cleaning, and Merging -->

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(glue)
library(httr2)
library(gt)
library(scales)
library(ggrepel)
library(ggbump)
library(ggforce)
```

```{r}
#| label: load-functions
#| message: false
#| warning: false

load_atus_data <- function(file_base = c("resp", "rost", "act", "cps")){
    if(!dir.exists(file.path("data", "mp02"))){
        dir.create(file.path("data", "mp02"), showWarnings=FALSE, recursive=TRUE)
    }
    
    file_base <- match.arg(file_base)
    
    file_name_out <- file.path("data", 
                               "mp02", 
                               glue("atus{file_base}_0324.dat"))
        
    if(!file.exists(file_name_out)){
        
        url_end <- glue("atus{file_base}-0324.zip")
        
        temp_zip <- tempfile(fileext=".zip")
        temp_dir <- tools::file_path_sans_ext(temp_zip)
        
        request("https://www.bls.gov") |>
            req_url_path("tus", "datafiles", url_end) |>
            req_headers(`User-Agent` = "Mozilla/5.0 (Macintosh; Intel Mac OS X 10.15; rv:149.0) Gecko/20100101 Firefox/149.0") |>
            req_perform(temp_zip)
    
        unzip(temp_zip, exdir = temp_dir)
        
        temp_file <- file.path(temp_dir, basename(file_name_out))
        
        file.copy(temp_file, file_name_out)
    }
    
    file <- read_csv(file_name_out, show_col_types = FALSE)
    
    # The original ALL-CAPS BLS columns are dropped; only the renamed or newly
    # created (not all-caps) columns are kept.
    keep_new_columns <- function(df) select(df, !matches("^[A-Z0-9_]+$", ignore.case = FALSE))
    
    switch(file_base,
        cps = file |>
            rename(participant_id = TUCASEID) |>
            rename_with(tolower),
        resp = file |>
            rename(
                participant_id = TUCASEID,
                survey_weight  = TUFNWGTP,
                time_alone     = TRTALONE,
                survey_year    = TUYEAR
            ) |>
            mutate(
                n_children = TRCHILDNUM,
                # Household composition (BLS: TRYHHCHILD = age of youngest household
                # child under 18, TRSPPRES = 1 spouse present, TESPEMPNOT = 1 spouse
                # employed / 2 not employed)
                youngest_child_age = if_else(TRYHHCHILD >= 0, as.numeric(TRYHHCHILD), NA_real_),
                spouse_present     = TRSPPRES == 1,
                spouse_employed    = case_match(TESPEMPNOT, 1 ~ TRUE, 2 ~ FALSE, .default = NA),
                # BLS definition of TELFS (labor force status)
                employment_status = case_match(TELFS,
                    1 ~ "Employed - at work",
                    2 ~ "Employed - absent",
                    3 ~ "Unemployed - on layoff",
                    4 ~ "Unemployed - looking",
                    5 ~ "Not in labor force",
                    .default = NA_character_),
                # Full/part time is only defined for the employed
                work_time = case_match(TRDPFTPT,
                    1 ~ "Full time",
                    2 ~ "Part time",
                    .default = NA_character_),
                # TESCHENR covers high school AND college; TESCHLVL tells them apart
                enrolled_in_school  = TESCHENR == 1,
                enrolled_in_college = TESCHENR == 1 & TESCHLVL == 2,
                # Actual diary date (the CPS month is 2 to 5 months earlier)
                diary_date  = ymd(TUDIARYDATE),
                diary_month = month(diary_date)
            ) |>
            keep_new_columns(),
        rost = file |>
            rename(participant_id = TUCASEID) |>
            mutate(
                sex         = case_match(TESEX, 1 ~ "M", 2 ~ "F"),
                roster_line = TULINENO,
                roster_age  = TEAGE
            ) |>
            keep_new_columns(),
        act = file |>
            rename(participant_id = TUCASEID) |>
            mutate(
                activity_n_of_day = TUACTIVITY_N,
                time_spent_min    = TUACTDUR24,
                start_time        = TUSTARTTIM,
                stop_time         = TUSTOPTIME,
                level1_code       = paste0(TRTIER1P, "0000"),
                level2_code       = paste0(TRTIER2P, "00"),
                # Six-digit activity code as text, so it matches the lexicon whether
                # readr reads the column as text (leading zeros) or as a number
                level3_code       = sprintf("%06d", as.integer(TRCODEP)),
                location_code     = TEWHERE
            ) |>
            keep_new_columns()
    )
}

load_atus_activities <- function(){
    if(!dir.exists(file.path("data", "mp02"))){
        dir.create(file.path("data", "mp02"), showWarnings=FALSE, recursive=TRUE)
    }
    
    dest_file <- file.path("data", "mp02", "atus_activity_codes.csv")
    
    if(!file.exists(dest_file)){
        download.file(
            "https://michael-weylandt.com/STA9750/mini/atus_activity_codes.csv",
            quiet    = TRUE,
            destfile = dest_file,
            mode     = "wb"
        )
    }
    
    # Six-digit activity code as text (same format as the activity file)
    read_csv(dest_file, show_col_types = FALSE) |>
        mutate(code_level2 = sprintf("%06d", as.integer(code_level2)),
               code_level3 = sprintf("%06d", as.integer(code_level3)))
}
```

```{r}
#| label: load-data
#| message: false
#| warning: false

atus_resp  <- load_atus_data("resp")
atus_rost  <- load_atus_data("rost")
atus_act   <- load_atus_data("act")
atus_codes <- load_atus_activities()
atus_cps   <- load_atus_data("cps")

# The ATUS respondent is always person line 1 (TULINENO = 1) in the ATUS files,
# so that is the row to take from the ATUS-CPS file. (PULINENO is the CPS
# household line number, which is usually 1 for the householder instead.)
cps_demo <- atus_cps |>
    filter(tulineno == 1) |>
    select(
        participant_id,
        age          = prtage,
        married_code = pemaritl,
        fam_income   = hefaminc,
        race         = ptdtrace,
        education    = peeduca,
        occupation   = peio1ocd,
        sex          = pesex
    ) |>
    mutate(
        married = case_when(
            married_code %in% c(1, 2)  ~ TRUE,
            married_code %in% 3:6      ~ FALSE,
            .default = NA
        ),
        sex = case_match(sex, 1 ~ "M", 2 ~ "F"),
        fam_income = case_match(fam_income,
            1  ~ "Less than $5,000",
            2  ~ "$5,000 to $7,499",
            3  ~ "$7,500 to $9,999",
            4  ~ "$10,000 to $12,499",
            5  ~ "$12,500 to $14,999",
            6  ~ "$15,000 to $19,999",
            7  ~ "$20,000 to $24,999",
            8  ~ "$25,000 to $29,999",
            9  ~ "$30,000 to $34,999",
            10 ~ "$35,000 to $39,999",
            11 ~ "$40,000 to $49,999",
            12 ~ "$50,000 to $59,999",
            13 ~ "$60,000 to $74,999",
            14 ~ "$75,000 to $99,999",
            15 ~ "$100,000 to $149,999",
            16 ~ "$150,000 and over",
            .default = NA_character_),
        education = case_match(education,
            31 ~ "Less than 1st grade",
            32 ~ "1st-4th grade",
            33 ~ "5th or 6th grade",
            34 ~ "7th or 8th grade",
            35 ~ "9th grade",
            36 ~ "10th grade",
            37 ~ "11th grade",
            38 ~ "12th grade no diploma",
            39 ~ "High school diploma",
            40 ~ "Some college no degree",
            41 ~ "Associate's - vocational",
            42 ~ "Associate's - academic",
            43 ~ "Bachelor's degree",
            44 ~ "Master's degree",
            45 ~ "Professional degree",
            46 ~ "Doctoral degree",
            .default = NA_character_)
    )

stopifnot(!anyDuplicated(cps_demo$participant_id))

atus_resp <- atus_resp |>
    left_join(cps_demo, join_by(participant_id)) |>
    mutate(
        age_group = cut(age,
                        breaks = c(14, 24, 34, 44, 54, 64, 85),
                        labels = c("15-24", "25-34", "35-44",
                                   "45-54", "55-64", "65+")),
        college_age_group = cut(age,
                                breaks = c(17, 22, 27, 32, 40, 85),
                                labels = c("18-22", "23-27", "28-32",
                                           "33-40", "41+")),
        # The ATUS has no "retired" flag: ages 65+ who are out of the labor
        # force is the proxy used throughout this report.
        work_group = case_when(
            employment_status %in% c("Employed - at work", "Employed - absent") &
                work_time == "Full time" ~ "Employed full time",
            employment_status == "Not in labor force" & age >= 65 ~
                "Retired (65+, not in labor force)",
            .default = NA_character_
        )
    )
```

```{r}
#| label: validate-demographics
#| message: false
#| warning: false

# Safety check: the age taken from the ATUS-CPS file must agree (within one
# year, since the CPS interview came 2 to 5 months earlier) with the age on the
# ATUS roster for the respondent. If not, the wrong person's row was linked.
roster_check <- atus_rost |>
    filter(roster_line == 1) |>
    select(participant_id, roster_age) |>
    inner_join(atus_resp |> select(participant_id, age), join_by(participant_id)) |>
    summarize(share = mean(abs(age - roster_age) <= 1, na.rm = TRUE)) |>
    pull(share)

stopifnot(
    "CPS demographics do not match the ATUS respondent: check the respondent row filter" =
        roster_check > 0.95
)
```

```{r}
#| label: activity-prep
#| message: false
#| warning: false

# ATUS lexicon categories used throughout. Everything is matched on the exact
# lexicon names (never on text fragments such as "care").
child_care_l2 <- c(
    "Caring for and Helping Household Children",
    "Activities Related to Household Children's Education",
    "Activities Related to Household Children's Health"
)

# The multi-year ATUS files contain combined activity codes (for example 180782,
# "travel related to shopping", or 020681, "pet care") that the course lexicon
# file does not list. The first four digits of any code are its tier 1 + tier 2
# category, so level 1 and level 2 names are matched on those four digits.
# Level 3 names are matched on the full six digits where the code is listed.
lexicon_l2 <- atus_codes |>
    mutate(code2 = substr(code_level2, 1, 4)) |>
    select(code2, task_level1, task_level2) |>
    distinct(code2, .keep_all = TRUE)
lexicon_l3 <- atus_codes |>
    select(code_level3, task_level3) |>
    distinct(code_level3, .keep_all = TRUE)

act_joined <- atus_act |>
    mutate(code2 = substr(level3_code, 1, 4)) |>
    left_join(lexicon_l3, join_by(level3_code == code_level3)) |>
    left_join(lexicon_l2, join_by(code2)) |>
    mutate(activity_label = case_when(
        task_level2 %in% child_care_l2                       ~ "Child Care",
        task_level1 == "Household Activities"                ~ "Household Activities",
        task_level1 %in% c("Socializing, Relaxing, and Leisure",
                           "Sports, Exercise, and Recreation") ~ "Leisure & Recreation",
        task_level1 == "Education"                           ~ "Education & Studying",
        task_level1 == "Personal Care"                       ~ "Sleep & Personal Care",
        task_level1 == "Work and Work-Related Activities"    ~ "Working",
        .default = NA_character_
    ))

# Share of diary minutes assigned to an activity category
matched_share <- sum(act_joined$time_spent_min[!is.na(act_joined$task_level1)], na.rm = TRUE) /
    sum(act_joined$time_spent_min, na.rm = TRUE)

# Hours per respondent on one diary day for the activities selected by `keep`.
# Respondents who did not do the activity get 0, so means are per respondent
# per day (not per logged activity).
person_hours <- function(keep) {
    act_joined |>
        filter({{ keep }}) |>
        group_by(participant_id) |>
        summarize(hours = sum(time_spent_min, na.rm = TRUE) / 60, .groups = "drop") |>
        right_join(atus_resp, join_by(participant_id)) |>
        mutate(hours = replace_na(hours, 0))
}

# Weighted quantile (so quartiles are consistent with the weighted means)
wtd_quantile <- function(x, w, p) {
    o  <- order(x)
    x  <- x[o]
    w  <- w[o]
    cw <- cumsum(w) / sum(w)
    x[which(cw >= p)[1]]
}

fmt1 <- function(x) formatC(x, format = "f", digits = 1)
fmt2 <- function(x) formatC(x, format = "f", digits = 2)
say  <- function(...) cat("\n", paste0(...), "\n\n", sep = "")
```

### How the Numbers Are Built

- **Unit of analysis.** Every number is an average per respondent per diary 
  day. Minutes are totaled for each respondent, and respondents who did not do 
  an activity count as zero. Averages use the ATUS final weights.
- **Primary activities only.** The ATUS diary records one main activity at a 
  time. Time spent on something while doing something else is not counted.
- **Activity groups** come from the exact ATUS lexicon names. "Child care" 
  means caring for household children, helping with their education, and 
  helping with their health.
- **Employment** follows the BLS definitions: employed (at work or absent), 
  unemployed (on layoff or looking), or not in the labor force. The ATUS does 
  not flag retirees or stay-at-home parents, so "retired" means ages 65 and 
  older who are not in the labor force. "Stay-at-home dad" is also a proxy, 
  defined with its profile table. Both are proxies, not labels from the survey.
- **Demographics** come from the ATUS-CPS file, using the respondent's row. 
  `r percent(roster_check, accuracy = 0.1)` of respondents' ages there match 
  their ATUS roster age within one year. `r percent(matched_share, accuracy = 0.1)` 
  of diary minutes are assigned to an activity category; the multi-year files 
  include combined codes (such as travel related to shopping) that are matched 
  to their category using the first four digits of the activity code.
- **Sample sizes** are shown with the results. Small groups give noisy 
  estimates.

## Exploratory Analysis

### Key Statistics at a Glance

<!-- Task 4: Exploratory Questions - Inline Values -->

```{r}
#| label: task4-inline
#| message: false
#| warning: false

# Q1: How many unique respondents since 2003?
n_respondents <- n_distinct(atus_resp$participant_id)

# Q2: How many different sports and exercise activities does ATUS track?
# (the "participating" list; the watching list repeats the same sports)
n_sports <- atus_codes |>
    filter(task_level2 == "Participating in Sports, Exercise, or Recreation",
           !str_starts(task_level3, "Other")) |>
    nrow()

# Q3: Who is out of the labor force, and who is (proxy) retired?
pct_nilf <- weighted.mean(atus_resp$employment_status == "Not in labor force",
                          atus_resp$survey_weight, na.rm = TRUE)
pct_retired_proxy <- weighted.mean(
    atus_resp$employment_status == "Not in labor force" & atus_resp$age >= 65,
    atus_resp$survey_weight, na.rm = TRUE)

# Q4: Average hours of sleep per night (all respondents)
avg_sleep_hrs <- person_hours(task_level2 == "Sleeping") |>
    summarize(avg = weighted.mean(hours, survey_weight)) |>
    pull(avg)

# Q5: Average hours parents spend on child care (respondents with a child
# under 18 in the household)
avg_childcare_hrs <- person_hours(task_level2 %in% child_care_l2) |>
    filter(n_children >= 1) |>
    summarize(avg = weighted.mean(hours, survey_weight)) |>
    pull(avg)
```

The ATUS dataset is vast in scope. Since 2003, 
**`r format(n_respondents, big.mark=",")`** unique Americans have participated 
in the survey, providing a foundation for understanding how time-spent 
patterns have evolved over the last two decades.

When it comes to sports, ATUS is surprisingly granular. The survey tracks 
**`r n_sports`** distinct sports and exercise activities at the most detailed 
activity level, allowing analysts to distinguish between activities like golf, 
basketball, and yoga rather than lumping them all into a single "sports" 
category.

About **`r percent(pct_nilf, accuracy=0.1)`** of respondents are not in the 
labor force, a group that includes retirees but also students, homemakers, and 
people who are unable to work. Adults 65 and older who are not in the labor 
force, our proxy for retirees, make up 
**`r percent(pct_retired_proxy, accuracy=0.1)`** of the sample, reflecting the 
aging of the US population.

Sleep is a universal human activity, and the data confirms this. 
On average, Americans sleep **`r round(avg_sleep_hrs, 1)`** hours per night, a 
figure that is in line with general health recommendations.

Finally, the research finds what may be obvious to many: parenting is a 
substantial time commitment. Among Americans with at least one child under 18 
at home, the average time devoted to child care activities, including 
education involvement and healthcare, is 
**`r fmt2(avg_childcare_hrs)`** hours per day. This figure likely 
understates total parenting time because it counts only primary activities; 
supervising children while doing something else is not included.

### Time-Use Patterns by the Numbers

<!-- Task 5: Exploratory Questions - Table Answers -->

#### Lawn Care: Retired vs. the Full-Time Employed

"The grass is greener"... a common assumption is that retirees, with all of the 
extra time freed up from their careers, are able to dedicate it to activities 
that felt out of reach during their working years. One such activity is lawn care 
and gardening. The table below puts this assumption to the test by 
comparing average time spent on these activities between retired and 
full-time employed Americans. Both groups are averaged over every respondent, 
including those who did no lawn care on their diary day.

```{r}
#| label: task5-lawn
#| message: false
#| warning: false

lawn_care <- person_hours(task_level3 == "Lawn, garden, and houseplant care") |>
    filter(!is.na(work_group)) |>
    group_by(work_group) |>
    summarize(
        respondents     = n(),
        avg_min_per_day = weighted.mean(hours, survey_weight) * 60,
        pct_any         = weighted.mean(hours > 0, survey_weight),
        avg_min_if_any  = weighted.mean(hours[hours > 0], survey_weight[hours > 0]) * 60,
        .groups = "drop"
    ) |>
    mutate(avg_hrs_per_week = avg_min_per_day * 7 / 60)

lawn_care |>
    gt() |>
    cols_label(
        work_group       = "Group",
        respondents      = "Respondents",
        avg_min_per_day  = "Avg. Minutes/Day (All)",
        pct_any          = "% Doing Any",
        avg_min_if_any   = "Avg. Minutes (If Any)",
        avg_hrs_per_week = "Avg. Hours/Week (All)"
    ) |>
    fmt_number(columns = c(avg_min_per_day, avg_min_if_any), decimals = 1) |>
    fmt_number(columns = avg_hrs_per_week, decimals = 2) |>
    fmt_number(columns = respondents, decimals = 0, use_seps = TRUE) |>
    fmt_percent(columns = pct_any, decimals = 1) |>
    tab_header(
        title    = "Time Spent on Lawn Care & Gardening",
        subtitle = "Retired (proxy) vs. Full-Time Employed Americans"
    ) |>
    tab_source_note("Source: ATUS 2003–2024 Public Use Microdata")
```

```{r}
#| label: task5-lawn-text
#| echo: false
#| results: asis

ret <- lawn_care |> filter(work_group == "Retired (65+, not in labor force)")
emp <- lawn_care |> filter(work_group == "Employed full time")
lawn_ratio <- ret$avg_min_per_day / emp$avg_min_per_day

nums <- paste0("(", fmt1(ret$avg_min_per_day), " vs. ", fmt1(emp$avg_min_per_day),
               " minutes per day; ", percent(ret$pct_any, accuracy = 1), " vs. ",
               percent(emp$pct_any, accuracy = 1), " report any lawn or garden time)")

if (lawn_ratio >= 1.15) {
    say("Retirees do spend more time on lawn care and gardening than their full-time employed counterparts ", nums, ". The common belief holds up: some of the free time that retirement creates goes to the yard.")
} else if (lawn_ratio <= 0.87) {
    say("Contrary to the common belief, retired Americans spend less time on lawn care and gardening than their full-time employed counterparts ", nums, ". This suggests that while retirees may have more free time, they do not necessarily choose to allocate it to lawn care, perhaps preferring other leisure activities or simply enjoying their outdoor spaces without the need for maintenance.")
} else {
    say("Maybe we already have the greenest grass. Retired Americans spend about as much time on lawn care and gardening as their full-time employed counterparts ", nums, ". This suggests that while retirees may have more free time, they do not necessarily choose to allocate it to lawn care, perhaps preferring other leisure activities or simply enjoying their outdoor spaces without the need for maintenance.")
}
```

---

#### Social & Recreational Time: Married vs. Unmarried

Marriage is often associated with a shift in social priorities. The table below 
examines whether married and unmarried Americans actually differ in how much time 
they dedicate to social and recreational activities. Here, that means everything 
the ATUS groups under "Socializing, Relaxing, and Leisure" (which includes 
watching TV) and under "Sports, Exercise, and Recreation".

```{r}
#| label: task5-social
#| message: false
#| warning: false

social_time <- person_hours(activity_label == "Leisure & Recreation") |>
    filter(!is.na(married)) |>
    group_by(married) |>
    summarize(
        respondents     = n(),
        avg_hrs_per_day = weighted.mean(hours, survey_weight),
        .groups = "drop"
    ) |>
    mutate(
        married         = if_else(married, "Married", "Unmarried"),
        avg_min_per_day = avg_hrs_per_day * 60
    ) |>
    select(married, respondents, avg_min_per_day, avg_hrs_per_day)

social_time |>
    gt() |>
    cols_label(
        married         = "Marital Status",
        respondents     = "Respondents",
        avg_min_per_day = "Avg. Minutes/Day",
        avg_hrs_per_day = "Avg. Hours/Day"
    ) |>
    fmt_number(columns = avg_min_per_day, decimals = 1) |>
    fmt_number(columns = avg_hrs_per_day, decimals = 2) |>
    fmt_number(columns = respondents, decimals = 0, use_seps = TRUE) |>
    tab_header(
        title    = "Time Spent on Social & Recreational Activities",
        subtitle = "Married vs. Unmarried Americans"
    ) |>
    tab_source_note("Source: ATUS 2003–2024 Public Use Microdata")
```

```{r}
#| label: task5-social-text
#| echo: false
#| results: asis

m_hrs <- social_time$avg_hrs_per_day[social_time$married == "Married"]
u_hrs <- social_time$avg_hrs_per_day[social_time$married == "Unmarried"]
nums  <- paste0("(", fmt2(u_hrs), " vs. ", fmt2(m_hrs), " hours per day)")

if (u_hrs - m_hrs >= 0.1) {
    say("Unmarried Americans spend more time on social and recreational activities than married Americans ", nums, ". One possible reason is that married Americans are more likely to have children at home, which takes up a large share of their time. It may also reflect that unmarried people rely more heavily on social activities as a primary source of connection and entertainment. This table does not test either explanation, and the two groups also differ in age and employment.")
} else if (m_hrs - u_hrs >= 0.1) {
    say("Married Americans spend more time on social and recreational activities than unmarried Americans ", nums, ". This table does not explain why, and the two groups also differ in age and employment.")
} else {
    say("Married and unmarried Americans spend about the same amount of time on social and recreational activities ", nums, ". The two groups also differ in age and employment, which this table does not account for.")
}
```

---

#### Where High-Earning Americans Spend Their Time

Understanding how affluent Americans allocate their time offers insight into 
the lifestyle trade-offs that come with high-income careers. Here, 
"high-earning" is defined as households with annual family income of at least 
$100,000 (nominal dollars, not adjusted for inflation across the survey years). 
The table ranks their top eight activity categories by average daily time spent.

```{r}
#| label: task5-highearn
#| message: false
#| warning: false

high_resp    <- atus_resp |>
    filter(fam_income %in% c("$100,000 to $149,999", "$150,000 and over"))
high_total_w <- sum(high_resp$survey_weight)

high_all <- act_joined |>
    filter(task_level1 != "Data Codes") |>
    inner_join(high_resp |> select(participant_id, survey_weight),
               join_by(participant_id)) |>
    group_by(task_level1) |>
    summarize(avg_min_per_day = sum(time_spent_min * survey_weight) / high_total_w,
              .groups = "drop") |>
    mutate(avg_hrs_per_day = avg_min_per_day / 60) |>
    arrange(desc(avg_min_per_day))

high_earner_time <- high_all |> slice_head(n = 8)

high_earner_time |>
    gt() |>
    cols_label(
        task_level1     = "Activity Category",
        avg_min_per_day = "Avg. Minutes/Day",
        avg_hrs_per_day = "Avg. Hours/Day"
    ) |>
    fmt_number(columns = avg_min_per_day, decimals = 1) |>
    fmt_number(columns = avg_hrs_per_day, decimals = 2) |>
    tab_header(
        title    = "Where High-Earning Americans Spend Their Time",
        subtitle = "Households earning $100,000 or more annually"
    ) |>
    tab_source_note("Source: ATUS 2003–2024 Public Use Microdata")
```

```{r}
#| label: task5-highearn-text
#| echo: false
#| results: asis

hrs_of <- function(label) high_all$avg_hrs_per_day[high_all$task_level1 == label]
work_h    <- hrs_of("Work and Work-Related Activities")
leisure_h <- hrs_of("Socializing, Relaxing, and Leisure")

say("The largest category is ", tolower(high_all$task_level1[1]), " at ",
    fmt1(high_all$avg_hrs_per_day[1]), " hours per day, followed by ",
    tolower(high_all$task_level1[2]), " (", fmt1(high_all$avg_hrs_per_day[2]),
    ") and ", tolower(high_all$task_level1[3]), " (", fmt1(high_all$avg_hrs_per_day[3]), ").")

lw_cmp <- if (abs(leisure_h - work_h) < 0.5) {
    "about as much as they spend on work"
} else if (leisure_h > work_h) {
    "more than they spend on work"
} else {
    "less than they spend on work"
}

say("Work and work-related activities average ", fmt1(work_h), " hours per day. That is an average over every diary day, weekends included, and over every respondent in these households, including those who were not working, so it is lower than the length of a typical workday. High earners also average ", fmt1(leisure_h), " hours per day on socializing, relaxing, and leisure, ", lw_cmp, ".",
    if (leisure_h >= 0.75 * work_h) " High earners do not give up all of their free time for professional output." else "")
```

---

#### Activities on a Bus, Train, Subway, Boat, or Ferry

Americans who find themselves aboard a bus, train, subway, boat, or ferry occupy 
a unique moment where their physical movement is fixed but their attention is 
entirely their own. The table below identifies the ten most commonly reported 
activities among Americans while aboard these vehicles, other than the travel 
itself. The ATUS codes the ride as travel, so riding is excluded here.

```{r}
#| label: task5-transit
#| message: false
#| warning: false

# ATUS "where" codes: 15 = bus, 16 = subway/train, 18 = boat/ferry
transit_codes <- c(15, 16, 18)

transit_activities <- act_joined |>
    filter(location_code %in% transit_codes,
           !task_level1 %in% c("Traveling", "Data Codes"),
           !str_detect(tolower(coalesce(task_level3, "")), "boat")) |>
    inner_join(atus_resp |> select(participant_id, survey_weight),
               join_by(participant_id)) |>
    # Combined multi-year codes have no tier-3 name; fall back to the tier-2 name
    mutate(activity = coalesce(task_level3, task_level2)) |>
    group_by(activity) |>
    summarize(
        n_people = n_distinct(participant_id),
        avg_min  = weighted.mean(time_spent_min, survey_weight, na.rm = TRUE),
        .groups = "drop"
    ) |>
    arrange(desc(n_people)) |>
    slice_head(n = 10)

transit_activities |>
    gt() |>
    cols_label(
        activity    = "Activity",
        n_people    = "Number of Respondents",
        avg_min     = "Avg. Minutes"
    ) |>
    fmt_number(columns = avg_min, decimals = 1) |>
    fmt_number(columns = n_people, decimals = 0) |>
    tab_header(
      title    = "Most Common Activities on a Bus, Train, Subway, Boat, or Ferry",
      subtitle = "Primary activities other than travel, reported while aboard these vehicles"
    ) |>
    tab_source_note("Source: ATUS 2003–2024 Public Use Microdata")
```

```{r}
#| label: task5-transit-text
#| echo: false
#| results: asis

top3 <- transit_activities |> slice_head(n = 3)
work_aboard <- "Work, main job" %in% transit_activities$activity

say("Because the ATUS codes the ride itself as travel, this table cannot show how riders use their trip. It shows the other primary activities respondents logged while aboard. The most commonly reported is ",
    tolower(top3$activity[1]), " (", comma(top3$n_people[1]), " respondents, averaging ",
    fmt1(top3$avg_min[1]), " minutes), followed by ", tolower(top3$activity[2]),
    " (", comma(top3$n_people[2]), ") and ", tolower(top3$activity[3]), " (", comma(top3$n_people[3]), ").",
    if (work_aboard) " Work appears in the list, and many of those entries are likely drivers, conductors, and crew for whom being aboard is the job; the ATUS does not identify them." else "")

say("For our app, the takeaway is a limit rather than an opportunity: the ATUS cannot tell us whether transit time is a good moment to prompt users to log their activities. That would need to be tested with app usage data.")
```

---

#### Daily Hours on Favorite Activities by Age Group

The table below tracks four universally common activities:

- Sleep
- Eating and drinking
- Sports and exercise
- TV watching

We split the results across six age groups to reveal how priorities shift 
as Americans grow older. Each number is an average over every respondent in the 
age group, including people who did none of the activity on their diary day.

```{r}
#| label: task5-fav
#| message: false
#| warning: false

fav_activities <- bind_rows(
    person_hours(task_level2 == "Sleeping") |>
        mutate(activity = "Sleep"),
    person_hours(task_level2 == "Eating and Drinking") |>
        mutate(activity = "Eating & Drinking"),
    person_hours(task_level2 == "Participating in Sports, Exercise, or Recreation") |>
        mutate(activity = "Sports & Exercise"),
    person_hours(task_level3 == "Television and movies (not religious)") |>
        mutate(activity = "TV Watching")
) |>
    filter(!is.na(age_group)) |>
    group_by(activity, age_group) |>
    summarize(
        avg_hrs = round(weighted.mean(hours, survey_weight), 2),
        .groups = "drop"
    )

fav_activities |>
    pivot_wider(names_from = age_group, values_from = avg_hrs) |>
    gt() |>
    cols_label(activity = "Activity") |>
    tab_spanner(columns = -activity, label = "Age Group") |>
    tab_header(
        title    = "Daily Hours Spent on Selected Activities by Age Group",
        subtitle = "Sleep, Eating & Drinking, Sports & Exercise, and TV Watching"
    ) |>
    tab_source_note("Source: ATUS 2003–2024 Public Use Microdata")
```

```{r}
#| label: task5-fav-text
#| echo: false
#| results: asis

get_fav <- function(act, grp) {
    fav_activities$avg_hrs[fav_activities$activity == act & fav_activities$age_group == grp]
}
sleep_all <- fav_activities |> filter(activity == "Sleep")
sleep_mid <- sleep_all |> filter(age_group %in% c("35-44", "45-54")) |> pull(avg_hrs) |> min()
u_shape   <- get_fav("Sleep", "15-24") > sleep_mid && get_fav("Sleep", "65+") > sleep_mid

sports_all   <- fav_activities |> filter(activity == "Sports & Exercise")
sports_peak  <- sports_all$age_group[which.max(sports_all$avg_hrs)]
tv15 <- get_fav("TV Watching", "15-24")
tv65 <- get_fav("TV Watching", "65+")
eat15 <- get_fav("Eating & Drinking", "15-24")
eat65 <- get_fav("Eating & Drinking", "65+")

say("Sleep ", if (diff(range(sleep_all$avg_hrs)) < 0.8) "stays within a narrow band across the lifespan" else "varies across the lifespan",
    ", between ", fmt2(min(sleep_all$avg_hrs)), " and ", fmt2(max(sleep_all$avg_hrs)), " hours per day.",
    if (u_shape) " The youngest and oldest Americans sleep slightly more than those in the middle years, which likely reflects the demands of work and family that peak during those years." else "")

say("Eating and drinking time ", if (eat65 > eat15) "grows" else "shrinks",
    " with age, from ", fmt2(eat15), " hours among 15 to 24 year olds to ", fmt2(eat65),
    " hours among those 65 and older",
    if (eat65 > eat15) ", perhaps reflecting a slower, more social approach to meals in retirement." else ".",
    " Sports and exercise average ", fmt2(max(sports_all$avg_hrs)), " hours at their peak (",
    sports_peak, ") and ", fmt2(get_fav("Sports & Exercise", "65+")), " hours among those 65 and older.")

say("TV watching ", if (tv65 > tv15) "rises" else "falls", " from ", fmt2(tv15),
    " hours among the youngest group to ", fmt2(tv65), " hours among Americans 65 and older, a change of ",
    percent(abs(tv65 / tv15 - 1), accuracy = 1),
    if (tv65 > tv15) ", consistent with retirement freeing up time for leisure, although these snapshots cannot separate aging from generational differences." else ".")
```

### Visual Representation of Time

<!-- Task 6: Exploratory Questions - Graph Answers -->

#### Household Activities by Age

```{r}
#| label: task6-graph1
#| message: false
#| warning: false

household_by_age <- person_hours(task_level1 == "Household Activities") |>
    filter(!is.na(age_group)) |>
    group_by(age_group) |>
    summarize(
        avg_hrs = weighted.mean(hours, survey_weight),
        .groups = "drop"
    )

ggplot(household_by_age, aes(x = age_group, y = avg_hrs, group = 1)) +
    geom_line(color = "#0072B2", linewidth = 1.2) +
    geom_point(color = "#0072B2", size = 3) +
    scale_y_continuous(limits = c(0, NA)) +
    labs(
        title    = "Time Spent on Household Activities by Age Group",
        subtitle = "Average hours per day, all respondents in each age group",
        x        = "Age Group",
        y        = "Average Hours per Day",
        caption  = "Source: ATUS 2003–2024 Public Use Microdata"
    )
```

```{r}
#| label: task6-graph1-text
#| echo: false
#| results: asis

hh_peak     <- as.character(household_by_age$age_group[which.max(household_by_age$avg_hrs)])
hh_peak_hrs <- max(household_by_age$avg_hrs)
hh_low      <- as.character(household_by_age$age_group[which.min(household_by_age$avg_hrs)])
hh_low_hrs  <- min(household_by_age$avg_hrs)
hh_rises    <- all(diff(household_by_age$avg_hrs) > 0)

say("Time spent on household activities is highest among ", hh_peak, " (", fmt2(hh_peak_hrs),
    " hours per day) and lowest among ", hh_low, " (", fmt2(hh_low_hrs), ").",
    if (hh_rises) " It rises steadily with age, which likely reflects a shift away from paid work toward home-based responsibilities and, for retirees, having more hours available during the day to devote to these tasks." else " It does not rise steadily with age.")
```

---

#### Summer Activities for Education Workers

Do teachers really spend their summers at the beach? The chart compares a typical 
day for K-12 and other teachers in the summer (June to August) with a day in the 
school year (September to May). Both include weekends and holidays. Months come 
from the actual diary date, and teachers are identified from Census occupation 
codes 2300 to 2360, a range that was revised during the survey period, so we 
treat the results as indicative.

```{r}
#| label: task6-graph2
#| message: false
#| warning: false

teachers <- atus_resp |>
    filter(!is.na(occupation), occupation >= 2300, occupation <= 2360,
           employment_status %in% c("Employed - at work", "Employed - absent"),
           !is.na(diary_month)) |>
    mutate(period = if_else(diary_month %in% 6:8,
                            "Summer (June-August)", "School year (September-May)"))

teacher_w <- teachers |>
    group_by(period) |>
    summarize(w_total = sum(survey_weight), n = n(), .groups = "drop")

teacher_time <- act_joined |>
    filter(task_level1 != "Data Codes") |>
    inner_join(teachers |> select(participant_id, survey_weight, period),
               join_by(participant_id)) |>
    group_by(period, task_level1) |>
    summarize(w_minutes = sum(time_spent_min * survey_weight), .groups = "drop") |>
    left_join(teacher_w, join_by(period)) |>
    mutate(avg_hrs = w_minutes / w_total / 60)

top_teacher_cats <- teacher_time |>
    group_by(task_level1) |>
    summarize(m = mean(avg_hrs), .groups = "drop") |>
    slice_max(m, n = 6) |>
    pull(task_level1)

teacher_caption <- paste0(
    "n = ", paste0(teacher_w$n, " (", teacher_w$period, ")", collapse = "; "),
    "\nSource: ATUS 2003–2024 Public Use Microdata")

teacher_time |>
    filter(task_level1 %in% top_teacher_cats) |>
    mutate(task_level1 = str_wrap(task_level1, 28)) |>
    ggplot(aes(x = reorder(task_level1, avg_hrs), y = avg_hrs, fill = period)) +
    geom_col(position = "dodge") +
    coord_flip() +
    scale_fill_manual(values = c("Summer (June-August)" = "#956C3B",
                                 "School year (September-May)" = "#0072B2")) +
    labs(
        title    = "How Teachers' Days Differ in Summer vs. the School Year",
        subtitle = "Average hours per day on the six largest activity categories",
        x        = NULL,
        y        = "Average Hours per Day",
        fill     = NULL,
        caption  = teacher_caption
    ) +
    theme(legend.position = "bottom")
```

```{r}
#| label: task6-graph2-text
#| echo: false
#| results: asis

tw <- function(period, cat) {
    teacher_time$avg_hrs[teacher_time$period == period & teacher_time$task_level1 == cat]
}
work_s  <- tw("Summer (June-August)", "Work and Work-Related Activities")
work_y  <- tw("School year (September-May)", "Work and Work-Related Activities")
work_ratio <- work_s / work_y
n_s <- teacher_w$n[teacher_w$period == "Summer (June-August)"]
n_y <- teacher_w$n[teacher_w$period == "School year (September-May)"]

lead <- paste0("Work and work-related activities average ", fmt2(work_s), " hours per day in the summer and ",
               fmt2(work_y), " in the school year (", n_s, " and ", n_y, " teacher diaries).")

if (work_ratio <= 0.5) {
    say(lead, " Work drops sharply in summer, so teachers do get a real break, although the average is still above zero, which fits professional development, lesson planning, or second jobs.")
} else if (work_ratio <= 0.85) {
    say(lead, " Work falls in the summer but remains substantial, suggesting that many teachers use the summer months for professional development, lesson planning, or even second jobs rather than taking an extended break from work. This has implications for our app: educators may benefit from features that help them manage their time during the summer, rather than assuming they have more free time to log leisure activities.")
} else {
    say(lead, " Teachers' routines change little between summer and the school year, suggesting that many use the summer months for professional development, lesson planning, or even second jobs. This has implications for our app: educators may benefit from features that help them manage their time effectively during the summer, rather than assuming they have more free time to log leisure activities.")
}
if (min(n_s, n_y) < 150) say("With fewer than 150 diaries in at least one period, these averages are noisy.")
```

---

#### Educational Activity Time Among College Students by Age

```{r}
#| label: task6-graph3
#| message: false
#| warning: false

college_resp <- atus_resp |>
    filter(enrolled_in_college, !is.na(college_age_group))

college_w <- college_resp |>
    group_by(college_age_group) |>
    summarize(w_total = sum(survey_weight), n = n(), .groups = "drop")

college_edu <- act_joined |>
    filter(task_level1 == "Education") |>
    inner_join(college_resp |> select(participant_id, survey_weight, college_age_group),
               join_by(participant_id)) |>
    group_by(college_age_group, task_level2) |>
    summarize(w_minutes = sum(time_spent_min * survey_weight), .groups = "drop") |>
    left_join(college_w, join_by(college_age_group)) |>
    mutate(
        avg_hrs   = w_minutes / w_total / 60,
        age_label = paste0(college_age_group, "\n(n = ", comma(n), ")")
    )

age_label_levels <- college_w |>
    arrange(college_age_group) |>
    mutate(lbl = paste0(college_age_group, "\n(n = ", comma(n), ")")) |>
    pull(lbl)

college_edu <- college_edu |>
    mutate(age_label = factor(age_label, levels = age_label_levels))

college_totals <- college_edu |>
    group_by(college_age_group) |>
    summarize(total_hrs = sum(avg_hrs), n = first(n), .groups = "drop")

ggplot(college_edu, aes(x = age_label, y = avg_hrs, fill = task_level2)) +
    geom_col(position = "stack") +
    scale_fill_brewer(type = "qual", palette = 2, name = "Activity Type",
                      labels = function(x) str_wrap(x, width = 25)) +
    labs(
        title    = "Educational Activity Time by Age Among College Students",
        subtitle = "Average hours per day, all enrolled college students in each age group",
        x        = "Age Group",
        y        = "Average Hours per Day",
        caption  = "Source: ATUS 2003–2024 Public Use Microdata"
    ) +
    theme(
        legend.position  = "bottom",
        legend.text      = element_text(size = 7),
        legend.key.size  = unit(0.4, "cm"),
        plot.margin      = margin(5, 5, 20, 5)
    ) +
    guides(fill = guide_legend(nrow = 3))
```

```{r}
#| label: task6-graph3-text
#| echo: false
#| results: asis

edu_peak <- as.character(college_totals$college_age_group[which.max(college_totals$total_hrs)])
tot18 <- college_totals$total_hrs[college_totals$college_age_group == "18-22"]
tot_older <- weighted.mean(college_totals$total_hrs[college_totals$college_age_group != "18-22"],
                           college_totals$n[college_totals$college_age_group != "18-22"])

if (edu_peak == "18-22") {
    say("Younger college students (ages 18 to 22) spend more total time on educational activities than their older peers: ",
        fmt2(tot18), " hours per day, compared with ", fmt2(tot_older), " for students 23 and older. This may reflect that traditional-age students are more likely to be enrolled full-time, while older students often balance coursework with work and family obligations.")
} else {
    say("Total time on educational activities is highest among students aged ", edu_peak, " (",
        fmt2(max(college_totals$total_hrs)), " hours per day). Students aged 18 to 22 average ", fmt2(tot18),
        " hours per day, compared with ", fmt2(tot_older), " for students 23 and older.")
}
```

---

#### Relative Popularity of Sports by Age Group

```{r}
#| label: task6-graph4
#| message: false
#| warning: false

sports_person <- act_joined |>
    filter(task_level2 %in% c("Participating in Sports, Exercise, or Recreation",
                              "Attending Sports or Recreational Events"),
           !is.na(task_level3),
           !str_detect(task_level3, "^Other"),
           !str_detect(tolower(task_level3), "vehicle touring")) |>
    mutate(
        # Strip action prefix to group watching + participating together
        sport_name = str_remove(tolower(task_level3), 
                                "^watching |^playing |^doing |^participating in "),
        # Clean up a few names for display
        sport_name = case_match(sport_name,
            "skiing, ice skating, snowboarding"  ~ "Skiing/Skating",
            "climbing, spelunking, caving"        ~ "Climbing/Caving",
            "weightlifting or strength training"  ~ "Weightlifting",
            "using cardiovascular equipment"      ~ "Cardio Equipment",
            "working out, unspecified"            ~ "Working Out",
            "people working out, unspecified"     ~ "Working Out",
            "rodeo competitions"                  ~ "Rodeo",
            "equestrian sports"                   ~ "Equestrian",
            "water sports"                        ~ "Water Sports",
            "racquet sports"                      ~ "Racquet Sports",
            "martial arts"                        ~ "Martial Arts",
            .default = str_to_title(sport_name)
        )
    ) |>
    group_by(participant_id, sport_name) |>
    summarize(minutes = sum(time_spent_min, na.rm = TRUE), .groups = "drop")

# Average minutes per respondent per day: divide total weighted minutes by the
# total weight of ALL respondents in the age group (so non-participants count as 0)
age_totals <- atus_resp |>
    filter(!is.na(age_group)) |>
    group_by(age_group) |>
    summarize(w_total = sum(survey_weight), .groups = "drop")

sports_ranked <- sports_person |>
    inner_join(atus_resp |> select(participant_id, survey_weight, age_group),
               join_by(participant_id)) |>
    filter(!is.na(age_group)) |>
    group_by(age_group, sport_name) |>
    summarize(w_minutes = sum(minutes * survey_weight), .groups = "drop") |>
    left_join(age_totals, join_by(age_group)) |>
    mutate(avg_min = w_minutes / w_total) |>
    group_by(age_group) |>
    mutate(rank = rank(-avg_min, ties.method = "first")) |>
    ungroup()

sports_by_age <- sports_ranked |>
    filter(rank <= 6) |>
    group_by(sport_name) |>
    filter(n() >= 3) |>
    ungroup()

# Label each line where it last appears
sport_labels <- sports_by_age |>
    group_by(sport_name) |>
    slice_max(as.integer(age_group), n = 1, with_ties = FALSE) |>
    ungroup()

# Safe, distinct color palette
safe_colors <- c("#0072B2", "#D55E00", "#009E73", "#956C3B",
                 "#CC79A7", "#8B0000", "#4B0082", "#2F4F4F",
                 "#556B2F", "#8B4513", "#483D8B", "#8B7355")

ggplot(sports_by_age, aes(x = age_group, y = rank,
                           color = sport_name, group = sport_name)) +
    geom_bump(linewidth = 1) +
    geom_point(size = 3) +
    geom_label_repel(
        data         = sport_labels,
        aes(label    = sport_name),
        hjust        = 0,
        nudge_x      = 0.5,
        size         = 2.8,
        direction    = "y",
        segment.size = 0.3,
        box.padding  = 0.6,
        force        = 8,
        fill         = "white",
        label.size   = 0.2
    ) +
    scale_y_reverse(breaks = 1:6) +
    scale_color_manual(values = colorRampPalette(safe_colors)(12)) +
    coord_cartesian(clip = "off") +
    theme_bw() +
    theme(
        legend.position = "none",
        axis.text.x     = element_text(angle = 15, hjust = 1),
        plot.margin     = margin(5, 160, 5, 5)
    ) +
    labs(
        title    = "Sports and Exercise Ranked by Time Spent, by Age Group",
        subtitle = "Ranking by average minutes per person per day | Participating and watching combined",
        x        = "Age Group",
        y        = "Rank (1 = Most Time Spent)",
        caption  = "Source: ATUS 2003–2024 Public Use Microdata"
    )
```

```{r}
#| label: task6-graph4-text
#| echo: false
#| results: asis

top_by_age <- sports_ranked |> filter(rank == 1) |> arrange(age_group)
top_list   <- glue_collapse(
    glue("{top_by_age$age_group}: {tolower(top_by_age$sport_name)}"),
    sep = "; ", last = "; and ")
in_all <- sports_by_age |> count(sport_name) |> filter(n == 6) |> pull(sport_name)

say("The bump plot reveals how sports and exercise shift across the American lifespan. Each ranking uses average minutes per person per day, counting everyone in the age group, so it reflects both how many people do an activity and how long they spend on it. The top activity in each age group is: ",
    top_list, ".",
    if (length(in_all) > 0) paste0(" ", glue_collapse(in_all, sep = ", ", last = " and "), " rank",
                                   if (length(in_all) == 1) "s" else "",
                                   " in the top six for every age group.") else "")

base_rank <- sports_ranked |> filter(sport_name == "Baseball")
seasonal_absent <- setdiff(c("Hockey", "Skiing/Skating"), unique(sports_by_age$sport_name))
say("Because each ranking averages over every diary day of the year, seasonal sports are averaged over many off-season days, which pulls them down relative to year-round activities. ",
    if (nrow(base_rank) == 0) {
        "Baseball, often called America's pastime, does not appear in the data at all."
    } else if (min(base_rank$rank) > 6) {
        paste0("This helps explain why baseball, often called America's pastime, does not make the top six in any age group (its best rank is ", min(base_rank$rank), ")",
               if (length(seasonal_absent) > 0) paste0(", and why ", glue_collapse(tolower(seasonal_absent), sep = ", ", last = " and "), " ",
                                                       if (length(seasonal_absent) == 1) "does" else "do", " not appear either") else "",
               ".")
    } else {
        "Baseball, often called America's pastime, does make the top six in at least one age group."
    })
```

---

#### Parental Care Time Per Child by Family Size

```{r}
#| label: task6-graph5
#| message: false
#| warning: false

child_time <- person_hours(task_level2 %in% child_care_l2) |>
    filter(n_children >= 1) |>
    mutate(hours_per_child = hours / n_children)

childcare_by_family <- child_time |>
    filter(n_children <= 6) |>
    group_by(n_children) |>
    summarize(
        n                 = n(),
        total_hrs         = weighted.mean(hours, survey_weight),
        avg_hrs_per_child = weighted.mean(hours_per_child, survey_weight),
        .groups = "drop"
    )

childcare_by_family |>
    mutate(family_label = paste0(n_children, "\n(n = ", comma(n), ")")) |>
    select(family_label, `Total child care (hours per day)` = total_hrs,
           `Per child (hours per day)` = avg_hrs_per_child) |>
    pivot_longer(-family_label, names_to = "measure", values_to = "hours") |>
    mutate(measure = factor(measure, levels = c("Total child care (hours per day)",
                                                "Per child (hours per day)"))) |>
    ggplot(aes(x = family_label, y = hours)) +
    geom_col(fill = "#D55E00") +
    facet_wrap(~ measure) +
    labs(
        title    = "Parental Child Care Time by Family Size",
        subtitle = "Total care time rises with family size more slowly than the number of children",
        x        = "Number of Children in Household",
        y        = "Average Hours per Day",
        caption  = "Source: ATUS 2003–2024 Public Use Microdata"
    )
```

```{r}
#| label: task6-graph5-text
#| echo: false
#| results: asis

pc1   <- childcare_by_family$avg_hrs_per_child[childcare_by_family$n_children == 1]
tot1  <- childcare_by_family$total_hrs[childcare_by_family$n_children == 1]
three_plus <- child_time |>
    filter(n_children >= 3) |>
    summarize(per_child = weighted.mean(hours_per_child, survey_weight),
              total     = weighted.mean(hours, survey_weight))

if (pc1 > three_plus$per_child) {
    say("As family size increases, the time parents can devote to each individual child drops: parents with one child spend ",
        fmt2(pc1), " hours per child per day, compared with ", fmt2(three_plus$per_child),
        " for parents of three or more, or ", fmt1(pc1 / three_plus$per_child), " times as much. ",
        if (three_plus$total > tot1) paste0("Total child care time rises, from ", fmt2(tot1), " to ", fmt2(three_plus$total),
                                              " hours per day, but not in proportion to the number of children. ") else "",
        "This is consistent with parental attention being a finite resource that gets divided as families grow, the idea behind \"middle-child syndrome\". The data cannot show whether any particular child, such as a middle child, receives less than their siblings.")
} else {
    say("Per-child care time does not fall as family size grows in these data: parents with one child spend ",
        fmt2(pc1), " hours per child per day, compared with ", fmt2(three_plus$per_child), " for parents of three or more.")
}
```

## Final Insights and Deliverable

### Time-Use Profiles for Target Demographics

<!-- Task 7: Time Use Profiles for Target Demographics -->

For each target group, the table shows the weighted average and the middle half 
of respondents (Q1 to Q3), averaged over everyone in the group, including those 
who did none of the activity. Each target is either a minimum ("at least") or a 
limit ("at most"), set near what peers in the group actually do. The last 
column is the share of the group that meets it.

```{r}
#| label: profile-functions
#| message: false
#| warning: false

# `targets` is a tibble with one row per activity: activity (an activity_label),
# target_hrs, and direction ("at least", "at most", or "none" for no target)
build_profile <- function(group_df, targets) {
    activities <- targets$activity

    totals <- act_joined |>
        filter(activity_label %in% activities) |>
        inner_join(group_df |> select(participant_id), join_by(participant_id)) |>
        group_by(participant_id, activity_label) |>
        summarize(hrs = sum(time_spent_min, na.rm = TRUE) / 60, .groups = "drop")

    group_df |>
        select(participant_id, survey_weight) |>
        cross_join(targets |> rename(activity_label = activity)) |>
        left_join(totals, join_by(participant_id, activity_label)) |>
        mutate(
            hrs   = replace_na(hrs, 0),
            meets = case_when(
                direction == "at least" ~ hrs >= target_hrs,
                direction == "at most"  ~ hrs <= target_hrs,
                .default = NA
            )
        ) |>
        group_by(activity_label, target_hrs, direction) |>
        summarize(
            low         = wtd_quantile(hrs, survey_weight, 0.25),
            avg_hrs     = weighted.mean(hrs, survey_weight),
            high        = wtd_quantile(hrs, survey_weight, 0.75),
            pct_meeting = weighted.mean(meets, survey_weight),
            .groups = "drop"
        ) |>
        mutate(activity_label = factor(activity_label, levels = activities)) |>
        arrange(activity_label) |>
        mutate(
            activity  = as.character(activity_label),
            goal      = if_else(direction == "none", "None",
                                paste(direction, fmt1(target_hrs))),
            # Only a miss in the target's direction counts as a shortfall
            shortfall = case_when(
                direction == "at least" ~ pmax(target_hrs - avg_hrs, 0),
                direction == "at most"  ~ pmax(avg_hrs - target_hrs, 0),
                .default = NA_real_
            )
        ) |>
        select(activity, low, avg_hrs, high, target_hrs, direction, goal,
               pct_meeting, shortfall)
}

profile_gt <- function(tbl, title, subtitle) {
    tbl |>
        select(activity, low, avg_hrs, high, goal, pct_meeting) |>
        gt() |>
        cols_label(
            activity    = "Activity",
            low         = "Low (Q1 hrs)",
            avg_hrs     = "Average (hrs/day)",
            high        = "High (Q3 hrs)",
            goal        = "Target (hrs/day)",
            pct_meeting = "% Meeting Target"
        ) |>
        fmt_number(columns = c(low, avg_hrs, high), decimals = 1) |>
        fmt_percent(columns = pct_meeting, decimals = 0) |>
        sub_missing(columns = pct_meeting, missing_text = "n/a") |>
        tab_header(title = title, subtitle = subtitle) |>
        tab_source_note("Source: ATUS 2003–2024 Public Use Microdata") |>
        tab_footnote(
            footnote = "Quartiles are taken over everyone in the group, including people who did none of the activity that day, so Q3 can be 0.0 while the average is above 0 when most of the group did none.",
            locations = cells_column_labels(columns = high)
        ) |>
        tab_footnote(
            footnote = "Targets are suggested goals for the app's gamification system, based on product manager judgment. \"At least\" targets are minimums and \"at most\" targets are limits. Activities with no target show the group average only.",
            locations = cells_column_labels(columns = goal)
        )
}

# Plain-language summary of a profile table
profile_text <- function(tbl, who, n) {
    desc <- glue_collapse(
        glue("time on {tolower(tbl$activity)} averages {fmt1(tbl$avg_hrs)} hours",
             "{ifelse(tbl$direction == 'none', '', paste0(' (target: ', tbl$goal, ')'))}"),
        sep = ", ", last = ", and ")
    short <- tbl |> filter(!is.na(shortfall), shortfall > 0)
    gap_txt <- if (nrow(short) == 0) {
        "On average, the group already meets all of its targets."
    } else {
        big <- short |> slice_max(shortfall, n = 1, with_ties = FALSE)
        paste0("The largest shortfall against a target is for ", tolower(big$activity),
               ", where the average of ", fmt1(big$avg_hrs), " hours is ", fmt1(big$shortfall),
               " hours ", if (big$direction == "at least") "below" else "above",
               " the target (", big$goal, "), and ", percent(big$pct_meeting, accuracy = 1),
               " of the group meets it.")
    }
    paste0("Among ", who, " (n = ", comma(n), "), ", desc, ". ", gap_txt)
}
```

```{r}
#| label: task7-sahd
#| message: false
#| warning: false

# ── Demographic 1: Stay-at-home dads (proxy) ─────────────────────────────────
# The ATUS does not identify stay-at-home parents directly. Base proxy: married
# men under 65 with a child under 18 who are not in the labor force. Tighter
# proxy: also an employed spouse in the household and a youngest child under 13.
sahd_base <- atus_resp |>
    filter(sex == "M", married %in% TRUE, n_children >= 1,
           employment_status == "Not in labor force",
           age >= 18, age < 65)

sahd_tight <- sahd_base |>
    filter(spouse_present %in% TRUE, spouse_employed %in% TRUE,
           youngest_child_age < 13)

# Use the tighter definition unless it leaves too few respondents
use_tight <- nrow(sahd_tight) >= 300
sahd <- if (use_tight) sahd_tight else sahd_base
sahd_definition <- if (use_tight) {
    "married men under 65 with an employed spouse and a youngest child under 13 who are not in the labor force"
} else {
    "married men under 65 with a child under 18 who are not in the labor force"
}

sahd_targets <- tribble(
    ~activity,              ~target_hrs, ~direction,
    "Child Care",           2.0,         "at least",
    "Household Activities", 3.0,         "at least",
    "Leisure & Recreation", 5.0,         "at most"
)
sahd_profile <- build_profile(sahd, sahd_targets)

profile_gt(sahd_profile,
           title    = "Time Use Profile: Stay-at-Home Dads (Proxy)",
           subtitle = glue("{str_to_sentence(sahd_definition)} (n = {comma(nrow(sahd))})"))
```

```{r}
#| label: task7-sahd-text
#| echo: false
#| results: asis

say("Stay-at-home dads represent a relatively small demographic in the United States. The ATUS does not identify them directly, so this group is approximated as ",
    sahd_definition, "; it also includes men who are out of the labor force for other reasons, such as disability or school.",
    if (!use_tight) " A tighter definition (an employed spouse and a youngest child under 13) left too few respondents, so the broader group is used." else if (nrow(sahd) < 500) " The sample is modest, so these numbers are indicative." else "")
say(profile_text(sahd_profile, "this group", nrow(sahd)))
```

---

```{r}
#| label: task7-college
#| message: false
#| warning: false

# ── Demographic 2: College undergraduates ────────────────────────────────────
# Definition: Age 18-24, enrolled in college (not high school)
college_students <- atus_resp |>
    filter(age >= 18, age <= 24, enrolled_in_college)

# Working is not treated as a problem to fix, so it has no target
college_targets <- tribble(
    ~activity,               ~target_hrs, ~direction,
    "Education & Studying",  5.0,         "at least",
    "Leisure & Recreation",  3.0,         "at most",
    "Sleep & Personal Care", 8.0,         "at least",
    "Working",               NA_real_,    "none"
)
college_profile <- build_profile(college_students, college_targets)

profile_gt(college_profile,
           title    = "Time Use Profile: College Undergraduates",
           subtitle = glue("Enrolled students ages 18–24 (n = {comma(nrow(college_students))})"))
```

```{r}
#| label: task7-college-text
#| echo: false
#| results: asis

say("College undergraduates face a unique time allocation challenge: balancing academic demands with social life, work, and adequate rest.")
say(profile_text(college_profile, "enrolled college students ages 18 to 24", nrow(college_students)))
```

---

<!-- Chose this third category to represent myself in this analysis -->

```{r}
#| label: task7-custom
#| message: false
#| warning: false

# ── Demographic 3: Career-Changing Graduate Students ─────────────────────────
# Married adults 30-40, enrolled in college, no children at home, working
# part-time or not working (unemployed or out of the labor force)
career_changers <- atus_resp |>
    filter(age >= 30, age <= 40, enrolled_in_college, married %in% TRUE,
           n_children == 0,
           (employment_status %in% c("Employed - at work", "Employed - absent") &
                work_time == "Part time") |
               employment_status %in% c("Unemployed - on layoff",
                                        "Unemployed - looking",
                                        "Not in labor force"))

career_targets <- tribble(
    ~activity,               ~target_hrs, ~direction,
    "Education & Studying",  5.0,         "at least",
    "Working",               2.0,         "at most",
    "Leisure & Recreation",  3.0,         "at most",
    "Sleep & Personal Care", 8.0,         "at least",
    "Household Activities",  1.5,         "at most"
)
career_profile <- build_profile(career_changers, career_targets)

profile_gt(career_profile,
           title    = "Time Use Profile: Career-Changing Graduate Students",
           subtitle = glue("Married adults ages 30–40, enrolled in college, no children, working part-time or not at all (n = {comma(nrow(career_changers))})"))
```

```{r}
#| label: task7-custom-text
#| echo: false
#| results: asis

study_career  <- career_profile$avg_hrs[career_profile$activity == "Education & Studying"]
study_college <- college_profile$avg_hrs[college_profile$activity == "Education & Studying"]
career_short  <- career_profile |>
    filter(!is.na(shortfall), shortfall > 0) |>
    slice_max(shortfall, n = 1, with_ties = FALSE)

say("Career-changing graduate students occupy a uniquely demanding position: unlike traditional students, they bring years of professional experience into the classroom while simultaneously managing the responsibilities of adult life. This group is small in the ATUS (n = ",
    comma(nrow(career_changers)), "), so these numbers are indicative rather than precise.")
say(profile_text(career_profile, "career-changing graduate students", nrow(career_changers)))
say("Their average of ", fmt1(study_career), " hours per day on education is ",
    if (study_career < study_college - 0.1) "less than" else if (study_career > study_college + 0.1) "more than" else "about the same as",
    " the ", fmt1(study_college), " hours of traditional undergraduates. ",
    if (nrow(career_short) > 0) paste0("The largest shortfall, for ", tolower(career_short$activity),
        ", is the clearest opportunity for \"gamification\" to make a meaningful difference.") else "")
```

---

<!-- Extra Credit Activity Rings Visualization -->

```{r}
#| label: extra-credit-rings
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 7

# ── Data setup ────────────────────────────────────────────────────────────────
# "Peer Average" comes from the career-changer profile computed above;
# "My Week" is my own estimated week.
ring_order  <- c("Education & Studying", "Working", "Leisure & Recreation",
                 "Sleep & Personal Care", "Household Activities")
activities  <- c("Education &\nStudying", "Working &\nWork-Related", 
                 "Leisure &\nRecreation", "Sleep &\nPersonal Care", 
                 "Household\nActivities")

target_hrs  <- career_profile$target_hrs[match(ring_order, career_profile$activity)]
typical_hrs <- career_profile$avg_hrs[match(ring_order, career_profile$activity)]
my_hrs      <- c(8.0, 1.0, 2.0, 8.0, 3.0)

target_pct_typical <- pmin(typical_hrs / target_hrs, 1)
target_pct_mine    <- pmin(my_hrs / target_hrs, 1)

n <- length(activities)

rings_data <- data.frame(
    activity   = rep(factor(activities, levels = rev(activities)), 2),
    group      = factor(rep(c("Peer Average", "My Week"), each = n),
                        levels = c("Peer Average", "My Week")),
    pct        = c(target_pct_typical, target_pct_mine),
    actual_hrs = round(c(typical_hrs, my_hrs), 1),
    target_hrs = rep(target_hrs, 2),
    ring       = rep(seq(n, 1), 2)
)

ring_colors <- c(
    "Education &\nStudying"   = "#0072B2",
    "Working &\nWork-Related" = "#D55E00",
    "Leisure &\nRecreation"   = "#009E73",
    "Sleep &\nPersonal Care"  = "#956C3B",
    "Household\nActivities"   = "#CC79A7"
)

ggplot(rings_data) +
    # Background track
    geom_arc_bar(
        aes(x0 = 0, y0 = 0, r0 = ring - 0.4, r = ring,
            start = 0, end = 2 * pi),
        fill = "#EEEEEE", color = NA
    ) +
    # Filled arcs
    geom_arc_bar(
        aes(x0 = 0, y0 = 0, r0 = ring - 0.4, r = ring,
            start = -pi/2, end = -pi/2 + 2 * pi * pct,
            fill = activity),
        color = NA
    ) +
    # Hours label at the bottom of each ring
    geom_text(
        data = rings_data |> filter(group == "My Week"),
        aes(x = 0, y = -ring + 0.2, 
            label = paste0(actual_hrs, "h"),
            color = activity),
        size = 2.8, fontface = "bold", hjust = 0.5
    ) +
    scale_fill_manual(
        values = ring_colors,
        name   = NULL,
        labels = c(
            "Education &\nStudying"   = "📚 Education & Studying",
            "Working &\nWork-Related" = "💼 Working & Work-Related",
            "Leisure &\nRecreation"   = "🎯 Leisure & Recreation",
            "Sleep &\nPersonal Care"  = "😴 Sleep & Personal Care",
            "Household\nActivities"   = "🏠 Household Activities"
        )
    ) +
    scale_color_manual(values = ring_colors, guide = "none") +
    coord_equal(clip = "off") +
    facet_wrap(~group) +
    theme_void() +
    theme(
        strip.text      = element_text(size = 13, face = "bold",
                                       color = "#3B2F2F", margin = margin(b = 10)),
        plot.title      = element_text(size = 15, face = "bold",
                                       color = "#956C3B", hjust = 0.5,
                                       margin = margin(b = 5)),
        plot.subtitle   = element_text(size = 10, color = "#8B7355",
                                       hjust = 0.5, margin = margin(b = 15)),
        plot.caption    = element_text(size = 8, color = "#8B7355", hjust = 0.5),
        plot.margin     = margin(10, 10, 10, 10),
        legend.position = "bottom",
        legend.text     = element_text(size = 9, color = "#3B2F2F"),
        legend.key.size = unit(0.5, "cm"),
        legend.spacing.x = unit(0.8, "cm")
    ) +
    guides(fill = guide_legend(nrow = 1)) +
    labs(
        title    = "Activity Rings: Career-Changing Graduate Student",
        subtitle = "Each ring shows hours spent as a % of the target value (capped at 100%)",
        caption  = "Source: ATUS 2003–2024 | Peer Average = career-changer group average; My Week = personal time estimates\nFor \"at most\" targets (working, leisure, household), a full ring means the limit was reached."
    )
```

## Conclusion

```{r}
#| label: conclusion-text
#| echo: false
#| results: asis

college_study_target <- college_targets$target_hrs[college_targets$activity == "Education & Studying"]
study_gap_dir <- if (study_college < college_study_target) "below" else "at or above"

say("This analysis, of over two decades of ATUS microdata, reveals several interesting patterns in how Americans allocate their time. Household activity time is highest among Americans aged ",
    hh_peak, " (", fmt1(hh_peak_hrs), " hours per day). College undergraduates average ",
    fmt1(study_college), " hours per day on educational activities, ", study_gap_dir,
    " the ", fmt1(college_study_target), "-hour study target we set for the app.",
    if (pc1 > three_plus$per_child) paste0(" Parental attention per child falls from ", fmt1(pc1),
        " hours per day in one-child households to ", fmt1(three_plus$per_child),
        " in households with three or more, consistent with attention being a finite resource that is divided as families grow.") else "")

say("For the time-tracking app, our three target demographics: stay-at-home dads, college undergraduates, and career-changing graduate students, each present distinct time-use patterns and distinct gaps between average and target behavior. These gaps represent the core opportunity for \"gamification\": nudging users toward healthier time allocation by showing them how they compare to peers and to evidence-based targets.")

say("Future analyses could incorporate longitudinal tracking to understand how individual time-use evolves over life stages, or integrate well-being survey data to identify which time allocations are most strongly associated with reported happiness and life satisfaction.")
```

::: {.callout-note title="AI Usage Statement" collapse="true"}

Claude (Anthropic) was used in this mini-project to assist with R code 
generation, debugging, and structuring the analytical pipeline. Specifically, 
AI assistance was used to help write and troubleshoot the data loading 
functions, join logic between ATUS files, and ggplot2 visualization code. 
The original written narrative, including the introduction, interpretive 
paragraphs, and conclusion, was written by the student without AI 
assistance, in accordance with course policy.

For the October 2026 republication, Claude was also used to audit the code, 
to implement the corrections described in the republication note, and to revise 
the narrative sentences affected by the corrected results.

:::

------------------------------------------------------------------------

This work ©2026 by Git User: raulsolanavarro was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at [MP #02](https://michael-weylandt.com/STA9750/mini/mini02.html)
