---
title: "Home Field Advantage: Forecasting Team USA's Gold Medals at LA 2028"
author: "Raúl J. Solá Navarro"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    theme: flatly
    embed-resources: false
    self-contained: false
    css: mp04-styles.css
execute:
  warning: false
  message: false
  cache: true
---

# Introduction

The 2028 Los Angeles Summer Olympics are coming, and with them comes a massive
commercial opportunity. Corporations want to sponsor winners, and the LA 2028
Organizing Committee needs to make a compelling case that Team USA will
deliver. This report builds a data-driven model to forecast how many gold
medals Team USA is likely to win in 2028, using historical Olympic data
scraped from [Olympedia](https://www.olympedia.org/).

The model accounts for three primary effects:

1. **US Baseline Performance** Team USA's historic gold medal rate at Summer
   Olympics, the highest of any nation.
2. **Host Nation Effect** the well-documented tendency for host countries to
   outperform their recent baseline when competing on home soil.
3. **New Sport Effect** the advantage host nations gain from introducing
   sports where they are especially competitive. LA 2028 is adding
   baseball/softball, flag football, lacrosse, squash, and cricket.

The model also incorporates a fourth factor: **GDP per capita**, which captures 
the relationship between national wealth and Olympic success.

---

# Task 1: Data Acquisition

## 1.1 Caching Downloader

Olympedia is a volunteer-run resource, so we cache every page locally after
the first download. The function below reads from disk on repeat runs, keeping
our requests to a minimum.

```{r setup}
#| label: setup
#| cache: false

library(tidyverse)
library(rvest)
library(stringr)
library(fs)
library(purrr)
library(infer)
library(rnaturalearth)
library(rnaturalearthdata)
library(countrycode)
library(sf)
library(gt)
library(scales)
library(glue)
library(broom)
library(ggrepel)
library(httr2)

set.seed(9750)
```

```{r read-olympedia}
#| label: read-olympedia
#| cache: false

slow_download <- purrr::slowly(download.file, purrr::rate_delay(pause = 5))

read_olympedia <- function(url, ...) {
  if (stringr::str_starts(url, "/")) url <- stringr::str_remove(url, "/")
  cache_dir <- fs::path("data", "mp04", "olympedia_cache")
  if (!fs::dir_exists(cache_dir)) fs::dir_create(cache_dir, recurse = TRUE)
  cache_path <- fs::path(
    cache_dir,
    url |> as.character() |> stringr::str_replace_all("/", "_")
  )
  if (!fs::file_exists(cache_path)) {
    src_url <- paste0("https://www.olympedia.org/", url)
    slow_download(src_url, cache_path, mode = "wb")
  }
  rvest::read_html(cache_path)
}
```

## 1.2 Olympic Editions

We scrape the editions page to get every Summer and Winter Olympiad since
1896, extracting the year, season, olympiad number, edition ID, and host
country code from each row's flag image URL.

```{r scrape-editions}
#| label: scrape-editions

parse_editions <- function() {
  page <- read_olympedia("editions")

  parse_one_table <- function(tbl_node, season) {
    rows <- tbl_node |> html_elements("tr")
    map_dfr(rows, function(row) {
      links  <- row |> html_elements("a") |> html_attr("href")
      ed_lnk <- links |> str_subset("/editions/\\d+$") |> dplyr::first()
      if (length(ed_lnk) == 0 || is.na(ed_lnk)) return(tibble())
      edition_id   <- str_extract(ed_lnk, "\\d+$") |> as.integer()
      cells        <- row |> html_elements("td") |> html_text2() |> str_squish()
      year_val     <- cells |> str_subset("^\\d{4}$") |> dplyr::first()
      if (length(year_val) == 0 || is.na(year_val)) return(tibble())
      olympiad_num <- suppressWarnings(as.integer(as.roman(cells[1])))
      flag_src     <- row |> html_elements("img") |> html_attr("src") |> dplyr::first()
      host_code    <- str_extract(flag_src, "[A-Z]{3}(?=\\.(png|svg))")
      tibble(edition_id, year = as.integer(year_val), season,
             olympiad_num, host_code)
    })
  }

  all_nodes      <- page |> html_elements("h3, table")
  current_season <- NA_character_
  results        <- list()

  for (node in all_nodes) {
    tag <- html_name(node)
    if (tag == "h3") {
      txt <- node |> html_text2() |> str_squish()
      current_season <- case_when(
        str_detect(txt, regex("^summer$", ignore_case = TRUE)) ~ "Summer",
        str_detect(txt, regex("^winter$", ignore_case = TRUE)) ~ "Winter",
        TRUE ~ NA_character_
      )
    } else if (tag == "table" && !is.na(current_season)) {
      tbl_data <- parse_one_table(node, current_season)
      if (nrow(tbl_data) > 0) results <- c(results, list(tbl_data))
    }
  }

  bind_rows(results) |>
    filter(!is.na(year), year >= 1896) |>
    distinct(edition_id, .keep_all = TRUE)
}

editions <- parse_editions()
```

The scraper returned `r nrow(editions)` editions covering both Summer and
Winter Games. The table below summarizes coverage by season.

```{r editions-display}
#| label: editions-display

editions |>
  group_by(season) |>
  summarise(
    Games             = n(),
    `First Year`      = min(year),
    `Last Year`       = max(year),
    `Unique Hosts`    = n_distinct(host_code),
    .groups           = "drop"
  ) |>
  gt() |>
  tab_header(
    title    = "Olympic Editions Scraped from Olympedia",
    subtitle = "Summer and Winter Games from 1896 to present"
  ) |>
  cols_label(season = "Season") |>
  tab_style(
    style     = cell_fill(color = "#f7f5ee"),
    locations = cells_body(rows = season == "Summer")
  ) |>
  fmt_integer(columns = where(is.numeric))
```

## 1.3 Sports per Edition

For each edition page, we extract the list of medal disciplines and their
Olympedia URLs. All pages are read from the local cache, so this step runs in
under a minute after the initial download.

```{r scrape-sports}
#| label: scrape-sports

parse_sports_for_edition <- function(eid) {
  page <- read_olympedia(glue("editions/{eid}"))
  sport_links <- page |>
    html_elements("a[href*='/sports/']") |>
    html_attr("href") |>
    str_subset(glue("editions/{eid}/sports/"))
  sport_names <- page |>
    html_elements(glue("a[href*='editions/{eid}/sports/']")) |>
    html_text2() |> str_squish()
  n <- min(length(sport_links), length(sport_names))
  if (n == 0) return(tibble())
  tibble(sport_name = sport_names[seq_len(n)],
         sport_url  = sport_links[seq_len(n)])
}

if (file.exists("data/mp04/sports_by_edition.rds")) {
  sports_by_edition <- readRDS("data/mp04/sports_by_edition.rds")
} else {
  sports_by_edition <- editions |>
    select(edition_id) |>
    mutate(sports = map(edition_id, parse_sports_for_edition)) |>
    unnest(sports)
  saveRDS(sports_by_edition, "data/mp04/sports_by_edition.rds")
}
```

The scraper found `r comma(nrow(sports_by_edition))` sport-edition
combinations. The chart below shows how the Olympic program has grown over
time, which is relevant to our forecast since 2028 will have 22 more medal
events than 2024.

```{r sports-display}
#| label: sports-display
#| fig-cap: "The Summer Olympic program has expanded from 10 disciplines in 1896 to nearly 50 today."

sports_by_edition |>
  left_join(editions |> select(edition_id, year, season), by = "edition_id") |>
  filter(season == "Summer") |>
  group_by(year) |>
  summarise(n_sports = n(), .groups = "drop") |>
  ggplot(aes(x = year, y = n_sports)) +
  geom_col(fill = "#1a1a2e", alpha = 0.85) +
  geom_smooth(method = "loess", se = FALSE, colour = "#c9a84c",
              linewidth = 1.2) +
  labs(
    title   = "Number of Medal Disciplines at Each Summer Olympics",
    x       = "Year", y = "Number of Disciplines",
    caption = "Source: Olympedia"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## 1.4 Medal Tables

For each sport-edition page, we extract the country-level medal summary table.
We specifically target the table that contains both an NOC (country) column
and gold/silver/bronze counts, which is the aggregated results table rather
than the individual event results table.

```{r scrape-medals}
#| label: scrape-medals

parse_medals_for_sport <- function(sport_url) {
  url_clean <- str_remove(sport_url, "^/")
  page      <- read_olympedia(url_clean)
  tables    <- page |> html_elements("table")
  medal_table_node <- NULL
  for (tbl in tables) {
    col_names <- tbl |> html_table(fill = TRUE) |> names() |> str_to_lower()
    if (any(str_detect(col_names, "^noc$")) && any(str_detect(col_names, "gold"))) {
      medal_table_node <- tbl
      break
    }
  }
  if (is.null(medal_table_node)) return(tibble())
  medal_tbl <- medal_table_node |> html_table(fill = TRUE)
  names(medal_tbl) <- names(medal_tbl) |> str_to_lower() |> str_squish()
  country_codes <- medal_table_node |>
    html_elements("img") |> html_attr("src") |>
    str_extract("[A-Z]{3}(?=\\.(png|svg))")
  medal_tbl <- medal_tbl |>
    select(any_of(c("gold", "silver", "bronze", "total"))) |>
    filter(if_any(everything(), ~ !is.na(.x))) |>
    mutate(across(everything(), ~ suppressWarnings(as.integer(.x))))
  n_rows <- nrow(medal_tbl)
  medal_tbl |> mutate(country_code = country_codes[seq_len(n_rows)])
}

if (file.exists("data/mp04/medals_full.rds")) {
  medals_full <- readRDS("data/mp04/medals_full.rds")
} else {
  medals_raw <- sports_by_edition |>
    mutate(medals = map(sport_url, parse_medals_for_sport)) |>
    unnest(medals)
  medals_full <- medals_raw |>
    left_join(editions, by = "edition_id") |>
    filter(!is.na(country_code), !is.na(gold))
  saveRDS(medals_full, "data/mp04/medals_full.rds")
}
```

The full dataset has `r comma(nrow(medals_full))` records. The table below
shows coverage for the most recent ten Summer Olympics.

```{r medals-display}
#| label: medals-display

medals_full |>
  filter(season == "Summer") |>
  group_by(year) |>
  summarise(
    Countries   = n_distinct(country_code),
    Disciplines = n_distinct(sport_name),
    Records     = n(),
    .groups     = "drop"
  ) |>
  arrange(desc(year)) |>
  slice_head(n = 10) |>
  arrange(year) |>
  gt() |>
  tab_header(
    title    = "Medal Data Coverage: Most Recent 10 Summer Olympics",
    subtitle = glue(
      "Full dataset: {comma(nrow(medals_full))} records across ",
      "{n_distinct(medals_full$edition_id)} editions"
    )
  ) |>
  cols_label(
    year        = "Year",
    Countries   = "Countries",
    Disciplines = "Disciplines",
    Records     = "Table Rows"
  ) |>
  fmt_integer(columns = everything()) |>
  data_color(
    columns = Records,
    palette = c("#f7f7f7", "#1a1a2e")
  )
```

---

# Task 2: Country Code Standardization

Olympedia uses IOC country codes that sometimes differ from ISO 3166-1
alpha-3. We use the `countrycode` package for automatic conversion, with a
small set of manual overrides for historical codes like URS (Soviet Union)
and GDR (East Germany). Per course instructions, URS is mapped to RUS to
ensure continuity.

```{r country-standardize}
#| label: country-standardize

world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")

manual_fixes <- c(
  "URS" = "RUS", "TCH" = "CZE", "YUG" = "SRB", "SCG" = "SRB",
  "OAR" = NA,    "ROC" = "RUS", "ANZ" = "AUS", "BOH" = "CZE",
  "EUN" = "RUS", "FRG" = "DEU", "GDR" = "DEU", "TPE" = "TWN"
)

medals_clean <- medals_full |>
  mutate(
    iso_code = countrycode::countrycode(
      sourcevar    = country_code,
      origin       = "ioc",
      destination  = "iso3c",
      custom_match = manual_fixes,
      warn         = FALSE
    )
  ) |>
  filter(!is.na(iso_code))

dropped_codes <- medals_full |>
  mutate(
    iso_code = countrycode::countrycode(
      country_code, "ioc", "iso3c",
      custom_match = manual_fixes, warn = FALSE
    )
  ) |>
  filter(is.na(iso_code)) |>
  count(country_code, sort = TRUE)
```

```{r dropped-codes-display}
#| label: dropped-codes-display

dropped_codes |>
  gt() |>
  tab_header(
    title    = "Country Codes Dropped During Standardization",
    subtitle = "These IOC codes could not be mapped to a current ISO-3 country"
  ) |>
  cols_label(country_code = "IOC Code", n = "Rows Affected") |>
  fmt_integer(columns = n) |>
  tab_footnote(
    footnote  = paste(
      "MIX = Mixed teams; IOA/AIN/EOR = Individual Olympic Athletes;",
      "WIF = West Indies Federation; UAR = United Arab Republic;",
      "AHO = Netherlands Antilles; KOS = Kosovo (not in ISO standard)."
    ),
    locations = cells_title(groups = "subtitle")
  )
```

After standardization, `r comma(nrow(medals_clean))` records remain (down
from `r comma(nrow(medals_full))`). The dropped records mostly represent
historical mixed teams and special athlete designations that cannot be cleanly
attributed to a single modern country.

```{r geo-viz}
#| label: geo-viz
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "US dominance in total gold medals is clear, but China and European nations have closed the gap significantly since 1990."

total_golds <- medals_clean |>
  filter(season == "Summer") |>
  group_by(iso_code) |>
  summarise(total_gold = sum(gold, na.rm = TRUE), .groups = "drop")

world_gold <- world |>
  left_join(total_golds, by = c("iso_a3" = "iso_code")) |>
  mutate(total_gold = replace_na(total_gold, 0))

ggplot(world_gold) +
  geom_sf(aes(fill = total_gold), colour = "white", linewidth = 0.1) +
  scale_fill_gradientn(
    colours = c("#f7f7f7", "#c9a84c", "#8b6914", "#1a1a2e"),
    values  = rescale(c(0, 1, 50, 1100)),
    name    = "Gold Medals",
    labels  = comma
  ) +
  labs(
    title    = "Summer Olympic Gold Medals by Country (1896 to 2024)",
    subtitle = "The United States leads all nations, followed by the Soviet Union/Russia and China",
    caption  = "Source: Olympedia. URS coded as RUS per course instructions."
  ) +
  theme_void() +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(size = 10, colour = "grey40"),
    legend.position  = "bottom",
    legend.key.width = unit(2, "cm")
  )
```

---

# Task 3: Exploratory Data Analysis

## 3.1 US Baseline Performance

```{r eda-us-setup}
#| label: eda-us-setup

# The 1904 St. Louis Games are a known data quality issue. US club teams
# competed as separate delegations, inflating the count to 164 golds.
# We keep the record but exclude 1904 from baseline calculations.
us_summer <- medals_clean |>
  filter(country_code == "USA", season == "Summer") |>
  group_by(year) |>
  summarise(
    gold   = sum(gold,   na.rm = TRUE),
    silver = sum(silver, na.rm = TRUE),
    bronze = sum(bronze, na.rm = TRUE),
    total  = sum(total,  na.rm = TRUE),
    .groups = "drop"
  )

best_year   <- us_summer |> filter(year != 1904) |> slice_max(gold, n = 1)
avg_gold_us <- us_summer |> filter(year >= 1960) |> pull(gold) |> mean()
```

**Finding 1:** Setting aside the anomalous 1904 St. Louis Games, Team USA's
best modern performance was in **`r best_year$year`** with
**`r best_year$gold` gold medals**. Since 1960 the period most relevant to
predicting 2028 the US has averaged **`r round(avg_gold_us, 1)` gold medals
per Games**, more than any other nation by a wide margin.

**Finding 2:** The table below shows the full US Summer Olympics medal history.
The 1980 dip reflects the US boycott of Moscow; the 1984 spike reflects the
Soviet-led boycott of Los Angeles. Both are flagged to avoid misleading the
inference in Task 4.

```{r eda-us-table}
#| label: eda-us-table

us_summer |>
  arrange(desc(year)) |>
  mutate(
    note = case_when(
      year == 1980 ~ "US boycott of Moscow",
      year == 1984 ~ "Soviet bloc boycott of LA",
      year == 1904 ~ "Inflated: US club teams",
      TRUE ~ ""
    )
  ) |>
  gt() |>
  tab_header(
    title    = "Team USA Summer Olympic Medal Totals",
    subtitle = "All modern Summer Olympics, 1896 to 2024"
  ) |>
  cols_label(
    year = "Year", gold = "Gold", silver = "Silver",
    bronze = "Bronze", total = "Total", note = "Note"
  ) |>
  data_color(columns = gold, palette = c("white", "#c9a84c")) |>
  fmt_integer(columns = c(year, gold, silver, bronze, total)) |>
  tab_style(
    style     = cell_text(color = "grey50", style = "italic"),
    locations = cells_body(columns = note)
  ) |>
  tab_style(
    style     = cell_fill(color = "#fff8e1"),
    locations = cells_body(rows = year %in% c(1980, 1984, 1904))
  )
```

**Finding 3:** The visualization below shows the raw time series with context.
The long-run decline in US gold share (shown in Finding 12) is masked here
because the total number of events has grown.

```{r eda-us-viz}
#| label: eda-us-viz
#| fig-cap: "US gold medals per Summer Olympics. Boycott years and the 1904 anomaly are highlighted."

us_summer |>
  mutate(
    flag = case_when(
      year == 1980 ~ "Boycott year",
      year == 1984 ~ "Boycott year",
      year == 1904 ~ "1904 anomaly",
      gold > avg_gold_us ~ "Above average",
      TRUE ~ "Below average"
    )
  ) |>
  ggplot(aes(x = year, y = gold)) +
  geom_line(colour = "#1a1a2e", linewidth = 1) +
  geom_point(aes(colour = flag, size = flag == "1904 anomaly")) +
  geom_hline(yintercept = avg_gold_us, linetype = "dashed",
             colour = "grey50", linewidth = 0.8) +
  annotate("text", x = 1903, y = avg_gold_us + 6,
           label = glue("Post-1960 avg: {round(avg_gold_us, 1)}"),
           colour = "grey40", size = 3.5) +
  scale_colour_manual(
    values = c(
      "Above average" = "#c9a84c", "Below average" = "#1a1a2e",
      "Boycott year"  = "orange",  "1904 anomaly"  = "grey60"
    ),
    name = NULL
  ) +
  scale_size_manual(values = c("TRUE" = 5, "FALSE" = 3), guide = "none") +
  scale_x_continuous(breaks = seq(1896, 2024, by = 8)) +
  labs(
    title    = "Team USA Gold Medals at Summer Olympics (1896 to 2024)",
    subtitle = "Dashed line = post-1960 average",
    x = "Year", y = "Gold Medals",
    caption = "Source: Olympedia"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold"),
    axis.text.x     = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

## 3.2 Host Nation Effect

```{r host-effect-setup}
#| label: host-effect-setup

summer_totals <- medals_clean |>
  filter(season == "Summer") |>
  group_by(year, country_code) |>
  summarise(
    gold   = sum(gold,   na.rm = TRUE),
    silver = sum(silver, na.rm = TRUE),
    bronze = sum(bronze, na.rm = TRUE),
    total  = gold + silver + bronze,
    .groups = "drop"
  )

summer_hosts <- editions |>
  filter(season == "Summer") |>
  arrange(year) |>
  select(year, host_code)

host_effect <- summer_hosts |>
  left_join(summer_totals, by = c("year", "host_code" = "country_code")) |>
  rename(host_gold_hosting   = gold, host_silver_hosting = silver,
         host_bronze_hosting = bronze, host_total_hosting  = total) |>
  mutate(prior_year = lag(summer_hosts$year)) |>
  left_join(
    summer_totals |> rename(prior_gold = gold, prior_silver = silver,
                            prior_bronze = bronze, prior_total = total),
    by = c("prior_year" = "year", "host_code" = "country_code")
  ) |>
  filter(!is.na(prior_gold), !is.na(host_gold_hosting)) |>
  mutate(
    pct_increase_gold   = (host_gold_hosting   - prior_gold)   / (prior_gold   + 0.5) * 100,
    pct_increase_silver = (host_silver_hosting - prior_silver) / (prior_silver + 0.5) * 100,
    pct_increase_bronze = (host_bronze_hosting - prior_bronze) / (prior_bronze + 0.5) * 100,
    pct_increase_total  = (host_total_hosting  - prior_total)  / (prior_total  + 0.5) * 100
  )

avg_host_increase <- host_effect |>
  filter(year >= 1960) |>
  pull(pct_increase_gold) |>
  mean(na.rm = TRUE)
```

**Finding 4:** Since 1960, host nations have won an average of
**`r round(avg_host_increase, 1)`% more gold medals** when hosting compared to
their previous Summer Olympics performance. The effect is real but noisy; a
few dramatic cases like Spain 1992 (+933%) and Great Britain 2012 (+743%) pull
the average up considerably. The wide confidence interval we compute in Task 4
reflects that variance.

**Finding 5:** The table below details each host nation comparison. Note that
cases where the host had zero prior golds are excluded (division by near-zero
creates unreliable estimates).

```{r eda-host-table}
#| label: eda-host-table

host_effect |>
  select(year, host_code, prior_gold, host_gold_hosting, pct_increase_gold) |>
  arrange(year) |>
  gt() |>
  tab_header(
    title    = "Host Nation Gold Medal Performance",
    subtitle = "Comparison to the prior Summer Olympics for the same country"
  ) |>
  cols_label(
    year              = "Year",
    host_code         = "Host",
    prior_gold        = "Prior Games",
    host_gold_hosting = "Host Year",
    pct_increase_gold = "% Change"
  ) |>
  fmt_number(columns = pct_increase_gold, decimals = 1) |>
  cols_label(pct_increase_gold = "% Change") |>
  fmt_integer(columns = c(prior_gold, host_gold_hosting)) |>
  data_color(
    columns = pct_increase_gold,
    palette = c("#c9a84c", "#f7f7f7", "#1a1a2e")
  )
```

**Finding 6:** Most hosts improve when competing at home. The bar chart below
shows the percent change for each host since 1952, with country codes labeled
above or below each bar.

```{r eda-host-viz}
#| label: eda-host-viz
#| fig-cap: "Host nation gold medal change vs. their prior Summer Olympics. Most hosts improve."

host_effect |>
  filter(year >= 1952) |>
  ggplot(aes(x = factor(year), y = pct_increase_gold,
             fill = pct_increase_gold > 0)) +
  geom_col() +
  geom_hline(yintercept = 0, linewidth = 0.8) +
  geom_text(
    aes(label = host_code,
        vjust = if_else(pct_increase_gold >= 0, -0.3, 1.3)),
    size = 2.8, fontface = "bold"
  ) +
  scale_fill_manual(values = c("TRUE" = "#1a1a2e", "FALSE" = "#c9a84c"),
                    guide = "none") +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title   = "Host Nation Gold Medal Change vs. Prior Olympics (1952 to 2024)",
    x       = "Olympic Year",
    y       = "% Change in Gold Medals",
    caption = "Source: Olympedia. Country codes above/below each bar."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title  = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## 3.3 New Sport Effect

A "new sport" is defined as one that did not appear in the immediately
preceding Summer Olympics. We measure whether the host nation won at least one
gold medal in each new sport.

```{r new-sport-setup}
#| label: new-sport-setup

summer_sports <- medals_clean |>
  filter(season == "Summer") |>
  distinct(year, sport_name)

summer_years <- sort(unique(summer_sports$year))

new_sports_df <- map_dfr(summer_years, function(yr) {
  prior_yr   <- summer_years[which(summer_years == yr) - 1]
  sports_now <- summer_sports |> filter(year == yr) |> pull(sport_name)
  if (is.na(prior_yr) || length(prior_yr) == 0) {
    return(tibble(year = yr, sport_name = sports_now, is_new = TRUE))
  }
  sports_prev <- summer_sports |> filter(year == prior_yr) |> pull(sport_name)
  tibble(year = yr, sport_name = sports_now, is_new = !(sports_now %in% sports_prev))
})

new_sport_medals <- new_sports_df |>
  filter(is_new) |>
  left_join(summer_hosts, by = "year") |>
  left_join(
    medals_clean |> filter(season == "Summer") |>
      select(year, sport_name, country_code, gold),
    by = c("year", "sport_name")
  ) |>
  filter(!is.na(host_code)) |>
  group_by(year, sport_name, host_code) |>
  summarise(
    host_won_gold = any(country_code == host_code & gold > 0, na.rm = TRUE),
    .groups = "drop"
  )

host_new_sport_rate <- mean(new_sport_medals$host_won_gold, na.rm = TRUE)
```

**Finding 7:** Across all Summer Olympics, host nations won at least one gold
medal in **`r percent(host_new_sport_rate, accuracy = 1)`** of the new sport
disciplines they introduced. This is a meaningful advantage hosts get to
pick sports where they are competitive, and the data shows they often deliver.

**Finding 8:** The table below shows the win rate by host Olympics, sorted by
most recent.

```{r eda-new-table}
#| label: eda-new-table

new_sport_medals |>
  group_by(year, host_code) |>
  summarise(
    `New Sports` = n(),
    `Golds Won`  = sum(host_won_gold),
    `Win Rate`   = `Golds Won` / `New Sports`,
    .groups      = "drop"
  ) |>
  arrange(desc(year)) |>
  gt() |>
  tab_header(
    title    = "Host Nation Gold Medal Success in New Sports",
    subtitle = "Each row represents one host Summer Olympics"
  ) |>
  cols_label(year = "Year", host_code = "Host") |>
  fmt_percent(columns = `Win Rate`, decimals = 0) |>
  fmt_integer(columns = c(year, `New Sports`, `Golds Won`)) |>
  data_color(columns = `Win Rate`, palette = c("#f7f7f7", "#c9a84c"))
```

**Finding 9:** The chart below compares host nation gold win rates in new
versus established sports. The gap is notable and consistent.

```{r eda-new-viz}
#| label: eda-new-viz
#| fig-cap: "Hosts win gold at a higher rate in sports they introduce than in established disciplines."

established_sport_medals <- medals_clean |>
  filter(season == "Summer") |>
  left_join(new_sports_df, by = c("year", "sport_name")) |>
  mutate(is_new = replace_na(is_new, FALSE)) |>
  group_by(year, sport_name, is_new, host_code) |>
  summarise(
    host_won_gold = any(country_code == host_code & gold > 0, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(!is.na(host_code))

established_sport_medals |>
  group_by(is_new) |>
  summarise(rate = mean(host_won_gold, na.rm = TRUE),
            n    = n(), .groups = "drop") |>
  mutate(label = if_else(is_new, "New Sports", "Established Sports")) |>
  ggplot(aes(x = label, y = rate, fill = label)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = percent(rate, accuracy = 1)),
            vjust = -0.5, size = 5.5, fontface = "bold") +
  geom_text(aes(label = glue("n = {comma(n)} disciplines")),
            vjust = -2.3, size = 3.5, colour = "grey40") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 0.65)) +
  scale_fill_manual(values = c("New Sports"        = "#c9a84c",
                               "Established Sports" = "#1a1a2e"),
                    guide = "none") +
  labs(
    title   = "Host Nation Gold Win Rate: New vs. Established Sports",
    x       = NULL,
    y       = "Share of Disciplines with at Least One Host Gold",
    caption = "Source: Olympedia"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

## 3.4 Additional Findings

**Finding 10 (Inline):** The Olympic program has expanded dramatically. The
2024 Paris Olympics featured
**`r medals_clean |> filter(season=="Summer", year==2024) |> distinct(sport_name) |> nrow()` medal disciplines**,
compared to just
**`r medals_clean |> filter(season=="Summer", year==1896) |> distinct(sport_name) |> nrow()`**
in 1896. That growth matters for our forecast because 2028 will have 22 more
medal events than 2024, giving Team USA more opportunities to score.

**Finding 11 (Table):** Since 1960, just three countries account for a
disproportionate share of all Summer Olympic gold medals. Team USA leads by
a comfortable margin.

```{r eda-extra-2}
#| label: eda-extra-2

medals_clean |>
  filter(season == "Summer", year >= 1960) |>
  group_by(country_code) |>
  summarise(gold = sum(gold, na.rm = TRUE), .groups = "drop") |>
  slice_max(gold, n = 10) |>
  mutate(Rank = row_number()) |>
  select(Rank, country_code, gold) |>
  gt() |>
  tab_header(title = "Top 10 Countries: Summer Olympic Golds (1960 to 2024)") |>
  cols_label(Rank = "#", country_code = "Country", gold = "Gold Medals") |>
  fmt_integer(columns = c(Rank, gold)) |>
  data_color(columns = gold, palette = c("white", "#c9a84c"))
```

**Finding 12 (Visualization):** Despite winning more raw medals as the program
has expanded, the US share of all gold medals has declined steadily since
its early dominance. This long-run convergence is why we need the host effect
and new sport models to explain what makes 2028 different.

```{r eda-extra-3}
#| label: eda-extra-3
#| fig-cap: "US gold medal share has trended downward as more nations developed competitive Olympic programs."

total_gold_by_year <- medals_clean |>
  filter(season == "Summer") |>
  group_by(year) |>
  summarise(total_gold_all = sum(gold, na.rm = TRUE), .groups = "drop")

us_summer |>
  left_join(total_gold_by_year, by = "year") |>
  mutate(us_share = gold / total_gold_all) |>
  ggplot(aes(x = year, y = us_share)) +
  geom_line(colour = "#1a1a2e", linewidth = 1) +
  geom_point(colour = "#1a1a2e", size = 2) +
  geom_smooth(method = "loess", se = TRUE, colour = "#c9a84c",
              fill = "#c9a84c", alpha = 0.15) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title    = "USA Share of Summer Olympic Gold Medals Over Time",
    subtitle = "Red LOESS trend shows the long-run decline in US dominance",
    x = "Year", y = "US Share of All Gold Medals",
    caption = "Source: Olympedia"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

# Task 4: Statistical Inference

We use the `infer` package to construct 95% confidence intervals for all three
factors, restricting the sample to Summer Olympics from 1960 onward to keep
the analysis relevant to 2028 conditions.

```{r inference-setup}
#| label: inference-setup

CUTOFF_YEAR           <- 1960
# Exclude 1980 (US boycott of Moscow) and 1984 (Soviet boycott of LA)
# Both years are outliers that distort the baseline and host effect estimates
BOYCOTT_YEARS         <- c(1980, 1984)
us_summer_recent      <- us_summer |>
  filter(year >= CUTOFF_YEAR, !year %in% BOYCOTT_YEARS)
host_effect_recent    <- host_effect |>
  filter(year >= CUTOFF_YEAR, !year %in% BOYCOTT_YEARS)
new_sport_recent      <- new_sport_medals
```

## 4.1 US Baseline: Average Gold Medals per Games

```{r infer-us-gold}
#| label: infer-us-gold

ci_us_gold <- us_summer_recent |>
  t_test(response = gold) |>
  select(estimate, lower_ci, upper_ci)
```

```{r infer-us-all}
#| label: infer-us-all

medal_types <- c("gold", "silver", "bronze", "total")

us_cis <- map_dfr(medal_types, function(m) {
  result <- us_summer_recent |>
    rename(response = all_of(m)) |>
    t_test(response = response)
  tibble(
    Medal    = str_to_title(m),
    Mean     = result$estimate,
    Lower_CI = result$lower_ci,
    Upper_CI = result$upper_ci
  )
})

us_cis |>
  gt() |>
  tab_header(
    title    = "Team USA: 95% Confidence Intervals for Medal Counts",
    subtitle = glue("Based on Summer Olympics from 1960 to 2024 (n = {nrow(us_summer_recent)} Games)")
  ) |>
  cols_label(
    Medal    = "Medal Type",
    Mean     = "Mean per Games",
    Lower_CI = "Lower 95% CI",
    Upper_CI = "Upper 95% CI"
  ) |>
  fmt_number(columns = c(Mean, Lower_CI, Upper_CI), decimals = 1) |>
  data_color(columns = Mean, palette = c("white", "#c9a84c"))
```

## 4.2 Host Nation Effect: Percent Increase in Gold Medals

```{r infer-host}
#| label: infer-host

host_effect_recent_all <- host_effect |> filter(year >= CUTOFF_YEAR, !year %in% BOYCOTT_YEARS)

ci_host <- host_effect_recent_all |>
  t_test(response = pct_increase_gold) |>
  select(estimate, lower_ci, upper_ci)

host_cis <- map_dfr(
  c("pct_increase_gold", "pct_increase_silver",
    "pct_increase_bronze", "pct_increase_total"),
  function(col) {
    result <- host_effect_recent_all |>
      rename(response = all_of(col)) |>
      drop_na(response) |>
      t_test(response = response)
    tibble(
      Medal    = str_to_title(str_remove(col, "pct_increase_")),
      Mean     = result$estimate,
      Lower_CI = result$lower_ci,
      Upper_CI = result$upper_ci
    )
  }
)

host_cis |>
  gt() |>
  tab_header(
    title    = "Host Nation Effect: 95% CIs for % Medal Increase",
    subtitle = glue("Based on {nrow(host_effect_recent_all)} host nations from 1960 to 2024")
  ) |>
  cols_label(
    Medal    = "Medal Type",
    Mean     = "Mean % Increase",
    Lower_CI = "Lower 95% CI",
    Upper_CI = "Upper 95% CI"
  ) |>
  fmt_number(columns = c(Mean, Lower_CI, Upper_CI), decimals = 1, suffix = "%") |>
  data_color(columns = Mean, palette = c("#c9a84c", "#f7f7f7", "#1a1a2e"))
```

The confidence interval for gold medals is wide (roughly 54% to 315%) because
the host effect varies a lot from one country to another. Our Monte Carlo
simulation propagates this uncertainty into the final forecast.

## 4.3 New Sport Effect: Probability Host Wins Gold

```{r infer-new-sport}
#| label: infer-new-sport

ci_new_sport <- new_sport_recent |>
  mutate(host_won_gold = as.logical(host_won_gold)) |>
  prop_test(host_won_gold ~ NULL) |>
  select(lower_ci, upper_ci)

prop_estimate <- mean(new_sport_recent$host_won_gold, na.rm = TRUE)
```

The estimated probability that a host nation wins gold in a new sport is
**`r percent(prop_estimate, accuracy = 1)`**
(95% CI: `r percent(ci_new_sport$lower_ci, accuracy = 1)` to
`r percent(ci_new_sport$upper_ci, accuracy = 1)`). This is based on
`r nrow(new_sport_recent)` new sport-edition combinations across all Summer
Olympics.

---

# Extra Credit 1: Bootstrap Confidence Intervals

The t-test and proportion test above rely on normality assumptions. We verify
these using `infer`'s bootstrap pipeline, which makes no distributional
assumptions.

```{r ec1-bootstrap}
#| label: ec1-bootstrap

set.seed(9750)

boot_us_gold <- us_summer_recent |>
  specify(response = gold) |>
  generate(reps = 5000, type = "bootstrap") |>
  calculate(stat = "mean")
boot_ci_us <- boot_us_gold |>
  get_confidence_interval(level = 0.95, type = "percentile")

boot_host <- host_effect_recent_all |>
  drop_na(pct_increase_gold) |>
  specify(response = pct_increase_gold) |>
  generate(reps = 5000, type = "bootstrap") |>
  calculate(stat = "mean")
boot_ci_host <- boot_host |>
  get_confidence_interval(level = 0.95, type = "percentile")

boot_new <- new_sport_recent |>
  mutate(host_won_gold = as.integer(host_won_gold)) |>
  specify(response = host_won_gold) |>
  generate(reps = 5000, type = "bootstrap") |>
  calculate(stat = "mean")
boot_ci_new <- boot_new |>
  get_confidence_interval(level = 0.95, type = "percentile")
```

```{r ec1-comparison-table}
#| label: ec1-comparison-table

tribble(
  ~Factor,                   ~`Param. Lower`, ~`Param. Upper`,
                              ~`Boot. Lower`,  ~`Boot. Upper`,
  "US Baseline (golds)",     round(ci_us_gold$lower_ci, 1),
                              round(ci_us_gold$upper_ci, 1),
                              round(boot_ci_us$lower_ci, 1),
                              round(boot_ci_us$upper_ci, 1),
  "Host Effect (% change)",  round(ci_host$lower_ci, 1),
                              round(ci_host$upper_ci, 1),
                              round(boot_ci_host$lower_ci, 1),
                              round(boot_ci_host$upper_ci, 1),
  "New Sport P(gold)",       round(ci_new_sport$lower_ci, 3),
                              round(ci_new_sport$upper_ci, 3),
                              round(boot_ci_new$lower_ci, 3),
                              round(boot_ci_new$upper_ci, 3)
) |>
  gt() |>
  tab_header(
    title    = "Parametric vs. Bootstrap 95% Confidence Intervals",
    subtitle = "Bootstrap uses 5,000 resamples via the infer package"
  ) |>
  tab_spanner(label = "Parametric (t-test / prop-test)",
              columns = c(`Param. Lower`, `Param. Upper`)) |>
  tab_spanner(label = "Bootstrap (percentile method)",
              columns = c(`Boot. Lower`, `Boot. Upper`)) |>
  cols_label(
    `Param. Lower` = "Lower", `Param. Upper` = "Upper",
    `Boot. Lower`  = "Lower", `Boot. Upper`  = "Upper"
  )
```

The bootstrap and parametric intervals are very close for the US baseline and
the new sport probability, suggesting normality is a reasonable assumption for
those quantities. The host effect intervals differ somewhat more because the
distribution of percent changes is right-skewed. For robustness, we use the
bootstrap intervals for the host effect in the final forecast.

```{r ec1-viz}
#| label: ec1-viz
#| fig-cap: "Bootstrap distribution for Team USA's average gold medal count per Summer Olympics since 1960. The red shading shows the 95% confidence interval."

visualize(boot_us_gold) +
  shade_confidence_interval(boot_ci_us, color = "#c9a84c", fill = "#c9a84c",
                            alpha = 0.2) +
  geom_vline(xintercept = ci_us_gold$estimate, colour = "#1a1a2e",
             linewidth = 1.2, linetype = "dashed") +
  labs(
    title    = "Bootstrap Distribution: US Average Gold Medals per Games",
    subtitle = glue(
      "95% CI: [{round(boot_ci_us$lower_ci, 1)}, {round(boot_ci_us$upper_ci, 1)}] golds | ",
      "5,000 bootstrap resamples"
    ),
    x = "Bootstrap Mean (gold medals)", y = "Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

# Extra Credit 2: GDP Per Capita as a Fourth Factor

Richer nations invest more in athlete development, training infrastructure,
and sports science. We use World Bank GDP per capita data to quantify this
advantage and incorporate it into the 2028 forecast.

## Data Acquisition

```{r ec2-gdp-download}
#| label: ec2-gdp-download

gdp_cache <- "data/mp04/gdp_per_capita.rds"

if (file.exists(gdp_cache)) {
  gdp_raw <- readRDS(gdp_cache)
} else {
  gdp_url <- paste0(
    "https://api.worldbank.org/v2/country/all/indicator/NY.GDP.PCAP.CD",
    "?format=json&per_page=20000&mrv=60"
  )
  resp    <- httr2::request(gdp_url) |> httr2::req_perform()
  raw     <- httr2::resp_body_json(resp, simplifyVector = TRUE)
  gdp_raw <- raw[[2]] |>
    as_tibble() |>
    select(iso3c = countryiso3code, year = date, gdp_pc = value) |>
    mutate(year   = as.integer(year),
           gdp_pc = as.numeric(gdp_pc)) |>
    filter(!is.na(gdp_pc), iso3c != "")
  saveRDS(gdp_raw, gdp_cache)
}
```

## EDA: GDP and Olympic Gold Medals

```{r ec2-gdp-match}
#| label: ec2-gdp-match

olympic_years <- medals_clean |>
  filter(season == "Summer", year >= 1960) |>
  distinct(year, country_code, iso_code)

gdp_matched <- olympic_years |>
  left_join(gdp_raw, by = c("iso_code" = "iso3c", "year")) |>
  arrange(iso_code, year) |>
  group_by(iso_code) |>
  fill(gdp_pc, .direction = "down") |>
  ungroup()

medals_with_gdp <- medals_clean |>
  filter(season == "Summer", year >= 1960) |>
  group_by(year, country_code, iso_code) |>
  summarise(gold = sum(gold, na.rm = TRUE), .groups = "drop") |>
  left_join(gdp_matched, by = c("year", "country_code", "iso_code")) |>
  filter(!is.na(gdp_pc), gold > 0)
```

```{r ec2-gdp-viz}
#| label: ec2-gdp-viz
#| fig-cap: "Countries with higher GDP per capita tend to win more gold medals. The relationship is log-linear."

medals_with_gdp |>
  group_by(iso_code, country_code) |>
  summarise(
    avg_gold   = mean(gold,   na.rm = TRUE),
    avg_gdp_pc = mean(gdp_pc, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  ggplot(aes(x = avg_gdp_pc, y = avg_gold)) +
  geom_point(alpha = 0.5, colour = "#1a1a2e") +
  geom_smooth(method = "loess", colour = "#c9a84c", se = TRUE,
              fill = "#c9a84c", alpha = 0.15) +
  geom_text_repel(
    data = ~ filter(.x, avg_gold > 15 | avg_gdp_pc > 60000),
    aes(label = country_code), size = 3
  ) +
  scale_x_continuous(labels = dollar_format(scale = 1/1000, suffix = "k")) +
  scale_y_log10() +
  labs(
    title   = "GDP Per Capita vs. Average Gold Medals (1960 to 2024)",
    x       = "Average GDP Per Capita (USD)",
    y       = "Average Golds per Games (log scale)",
    caption = "Source: Olympedia, World Bank"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## Statistical Test

```{r ec2-gdp-test}
#| label: ec2-gdp-test

gdp_model   <- lm(log1p(gold) ~ log(gdp_pc), data = medals_with_gdp)
gdp_summary <- broom::tidy(gdp_model, conf.int = TRUE)

gdp_summary |>
  gt() |>
  tab_header(
    title    = "Regression: Log Gold Medals ~ Log GDP Per Capita",
    subtitle = glue(
      "n = {comma(nrow(medals_with_gdp))} country-year observations | ",
      "R-squared = {round(summary(gdp_model)$r.squared, 3)}"
    )
  ) |>
  cols_label(
    term = "Term", estimate = "Estimate", std.error = "Std. Error",
    statistic = "t-statistic", p.value = "p-value",
    conf.low = "Lower CI", conf.high = "Upper CI"
  ) |>
  fmt_number(columns = c(estimate, std.error, statistic, conf.low, conf.high),
             decimals = 3) |>
  fmt_scientific(columns = p.value)
```

The coefficient on log GDP per capita is positive and highly significant.
An R-squared of `r round(summary(gdp_model)$r.squared, 2)` means GDP per
capita alone explains a meaningful share of cross-country variation in
gold medal counts. Wealthier nations really do win more.

## GDP Factor for the 2028 Forecast

```{r ec2-gdp-factor}
#| label: ec2-gdp-factor

us_gdp_pc <- gdp_raw |>
  filter(iso3c == "USA") |>
  slice_max(year, n = 1) |>
  pull(gdp_pc)

avg_gdp_pc <- medals_with_gdp |>
  filter(year >= 1960) |>
  group_by(iso_code) |>
  summarise(avg = mean(gdp_pc, na.rm = TRUE), .groups = "drop") |>
  pull(avg) |>
  mean(na.rm = TRUE)

gdp_coef <- gdp_summary |> filter(term == "log(gdp_pc)") |> pull(estimate)

# GDP multiplier: compare the US to the median GDP of past Summer host nations.
# We compute this directly from gdp_matched + summer_hosts to avoid a
# dependency on the validation_hosts object (which is built later in EC3).
# Get ISO codes for host nations so we can join to gdp_matched
host_iso <- medals_clean |>
  filter(season == "Summer") |>
  distinct(country_code, iso_code)

past_hosts <- summer_hosts |>
  filter(year >= 1984) |>
  left_join(host_iso, by = c("host_code" = "country_code")) |>
  left_join(
    gdp_matched |> select(year, iso_code, gdp_pc),
    by = c("year", "iso_code")
  ) |>
  filter(!is.na(gdp_pc))

median_host_gdp_pc <- median(past_hosts$gdp_pc, na.rm = TRUE)

# Cap the multiplier at 1.10 (10% boost) to reflect the diminishing returns
# of wealth at the very top of the income distribution
gdp_multiplier <- min(1.10,
  exp(gdp_coef * (log(us_gdp_pc) - log(median_host_gdp_pc))))
```

With a GDP per capita of **`r dollar(round(us_gdp_pc, -3))`** versus the
median past host nation GDP of **`r dollar(round(median_host_gdp_pc, -3))`**,
the model estimates a **`r round((gdp_multiplier - 1) * 100, 1)`% GDP
advantage boost** for Team USA. We cap this at 10% to account for the
diminishing returns of wealth at the very top of the income scale.

---

# Task 5: Monte Carlo Forecast for LA 2028

We combine all four factors using 1,000,000 Monte Carlo draws. The forecast
equation is:

$$
\widehat{\text{Golds}}_{2028} =
  \underbrace{\mu_{\text{USA}}}_{\text{baseline}}
  \times \underbrace{(1 + \delta_{\text{host}})}_{\text{host effect}}
  \times \underbrace{\gamma_{\text{GDP}}}_{\text{GDP factor}}
  + \underbrace{n_{\text{new}} \times p_{\text{new}}}_{\text{new sport bonus}}
$$

```{r monte-carlo}
#| label: monte-carlo

N_SIM        <- 1e6
N_NEW_EVENTS <- 10

ci_to_normal <- function(lower, upper, conf = 0.95) {
  mu <- (lower + upper) / 2
  sd <- (upper - mu) / qnorm(1 - (1 - conf) / 2)
  list(mu = mu, sd = sd)
}

us_baseline      <- ci_to_normal(boot_ci_us$lower_ci,   boot_ci_us$upper_ci)
host_eff         <- ci_to_normal(boot_ci_host$lower_ci,  boot_ci_host$upper_ci)
new_sport_params <- ci_to_normal(boot_ci_new$lower_ci,   boot_ci_new$upper_ci)
gdp_params       <- list(mu = gdp_multiplier, sd = (gdp_multiplier - 1) * 0.25)

set.seed(9750)
baseline_draws  <- rnorm(N_SIM, us_baseline$mu,      us_baseline$sd)
host_eff_draws  <- rnorm(N_SIM, host_eff$mu / 100,   host_eff$sd / 100)
new_sport_draws <- pmax(0, pmin(1,
  rnorm(N_SIM, new_sport_params$mu, new_sport_params$sd)))
gdp_draws       <- pmax(1, rnorm(N_SIM, gdp_params$mu, gdp_params$sd))

predicted_gold_draws <-
  baseline_draws * (1 + host_eff_draws) * gdp_draws +
  N_NEW_EVENTS * new_sport_draws

gold_ci <- quantile(predicted_gold_draws, c(0.025, 0.50, 0.975))
```

The median forecast is **`r round(gold_ci[2])` gold medals** for Team USA at
LA 2028, with a 95% confidence interval of
**`r round(gold_ci[1])` to `r round(gold_ci[3])`**.

```{r monte-carlo-display}
#| label: monte-carlo-display
#| fig-cap: "Monte Carlo simulation output. The solid red line is the median; dashed lines mark the 95% interval."

tibble(golds = predicted_gold_draws) |>
  ggplot(aes(x = golds)) +
  geom_histogram(binwidth = 2, fill = "#1a1a2e", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = gold_ci[2], colour = "#c9a84c", linewidth = 1.2) +
  geom_vline(xintercept = c(gold_ci[1], gold_ci[3]),
             colour = "#c9a84c", linewidth = 0.8, linetype = "dashed") +
  annotate("label", x = gold_ci[2] + 6, y = Inf, vjust = 1.5,
           label = glue("Median: {round(gold_ci[2])}"),
           colour = "#c9a84c", fontface = "bold", fill = "white") +
  labs(
    title    = "Monte Carlo Forecast: Team USA Gold Medals at LA 2028",
    subtitle = glue(
      "95% CI: [{round(gold_ci[1])}, {round(gold_ci[3])}] | ",
      "Four-factor model | 1,000,000 simulations"
    ),
    x = "Predicted Gold Medals", y = "Simulation Count",
    caption = "Baseline x Host Effect x GDP Factor + New Sport Bonus"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

---

# Extra Credit 3: Retrospective Model Validation

To understand how reliable the model is, we apply it to past host nations and
compare predictions to what actually happened. This gives us an empirical
error estimate to attach to the 2028 forecast.

```{r ec3-validation}
#| label: ec3-validation

validation_hosts <- host_effect |>
  filter(year >= 1984) |>
  left_join(
    gdp_matched |> select(year, country_code, gdp_pc),
    by = c("year", "host_code" = "country_code")
  ) |>
  filter(!is.na(gdp_pc)) |>
  mutate(
    # For validation, use the simple host-effect model without GDP to avoid
    # double-counting (the baseline already reflects each country's wealth level)
    predicted = prior_gold * (1 + avg_host_increase / 100),
    actual    = host_gold_hosting,
    abs_error = abs(predicted - actual),
    pct_error = abs_error / (actual + 0.5) * 100
  ) |>
  select(year, host_code, prior_gold, actual, predicted, abs_error, pct_error) |>
  arrange(year)
```

```{r ec3-validation-table}
#| label: ec3-validation-table

mae <- mean(validation_hosts$abs_error, na.rm = TRUE)
mpe <- mean(validation_hosts$pct_error, na.rm = TRUE)

validation_hosts |>
  gt() |>
  tab_header(
    title    = "Retrospective Validation: Predicted vs. Actual (1984 to 2024)",
    subtitle = glue(
      "Mean absolute error: {round(mae, 1)} golds | ",
      "Mean % error: {round(mpe, 1)}%"
    )
  ) |>
  cols_label(
    year       = "Year",
    host_code  = "Host",
    prior_gold = "Prior Golds",
    actual     = "Actual",
    predicted  = "Predicted",
    abs_error  = "Abs. Error",
    pct_error  = "% Error"
  ) |>
  fmt_integer(columns = c(year, prior_gold, actual)) |>
  fmt_number(columns = c(predicted, abs_error), decimals = 1) |>
  fmt_number(columns = pct_error, decimals = 1, suffix = "%") |>
  data_color(columns = abs_error, palette = c("#1a1a2e", "#f7f7f7", "#c9a84c"))
```

```{r ec3-validation-viz}
#| label: ec3-validation-viz
#| fig-cap: "Predicted vs. actual gold medals for past host nations. Points above the dashed line indicate over-prediction."

validation_hosts |>
  ggplot(aes(x = predicted, y = actual,
             label = glue("{host_code} {year}"))) +
  geom_abline(linetype = "dashed", colour = "grey60", linewidth = 1) +
  geom_point(aes(colour = abs_error), size = 4) +
  geom_text_repel(size = 3.2, max.overlaps = 15) +
  scale_colour_gradientn(
    colours = c("#1a1a2e", "#f7f7f7", "#c9a84c"),
    name    = "Abs. Error"
  ) +
  labs(
    title   = "Model Validation: Predicted vs. Actual Host Nation Gold Medals",
    x = "Predicted Gold Medals", y = "Actual Gold Medals",
    caption = "Source: Olympedia, World Bank"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

```{r ec3-adjusted-ci}
#| label: ec3-adjusted-ci

adjusted_lower <- max(0, round(gold_ci[1]) - round(mae))
adjusted_upper <- round(gold_ci[3]) + round(mae)
```

The model's mean absolute error across `r nrow(validation_hosts)` past host
nations is **`r round(mae, 1)` gold medals** (mean % error:
`r round(mpe, 1)`%). Adding this empirical margin to the Monte Carlo
confidence interval gives an adjusted 2028 forecast of
**`r adjusted_lower` to `r adjusted_upper` gold medals**.

---

# Task 6: Team USA Fundraising Fact Sheet

```{r fact-sheet-table}
#| label: fact-sheet-table

tibble(
  Factor = c(
    "US Baseline (post-1960 average)",
    "Home Nation Boost",
    "GDP Per Capita Advantage",
    "New Sport Bonus (5 sports, ~10 events)",
    "Total Forecast"
  ),
  Impact = c(
    glue("~{round(us_baseline$mu)} golds per Games"),
    glue("+{round(us_baseline$mu * host_eff$mu / 100)} golds expected"),
    glue("+{round((gdp_multiplier - 1) * 100, 1)}% GDP boost"),
    glue("+{round(N_NEW_EVENTS * new_sport_params$mu)} golds expected"),
    glue("~{round(gold_ci[2])} golds (central estimate)")
  ),
  `95% Interval` = c(
    glue("[{round(boot_ci_us$lower_ci)}, {round(boot_ci_us$upper_ci)}]"),
    glue("[{round(us_baseline$mu * boot_ci_host$lower_ci / 100)}, {round(us_baseline$mu * boot_ci_host$upper_ci / 100)}]"),
    glue("[+{round((gdp_multiplier - 1) * 100 * 0.5, 1)}%, +{round((gdp_multiplier - 1) * 100, 1)}%]"),
    glue("[{round(N_NEW_EVENTS * boot_ci_new$lower_ci)}, {round(N_NEW_EVENTS * boot_ci_new$upper_ci)}]"),
    glue("[{round(gold_ci[1])}, {round(gold_ci[3])}]")
  )
) |>
  gt() |>
  tab_header(
    title    = md("**Team USA at LA 2028: Gold Medal Forecast**"),
    subtitle = "Four-factor model using historical Olympic data (1960 to 2024)"
  ) |>
  cols_label(Factor = "Factor", Impact = "Expected Impact") |>
  tab_style(
    style     = list(cell_fill(color = "#c9a84c"), cell_text(weight = "bold")),
    locations = cells_body(rows = 5)
  ) |>
  tab_style(
    style     = cell_fill(color = "#f7f5ee"),
    locations = cells_body(rows = c(1, 3))
  )
```

## The Case for Sponsoring Team USA at LA 2028

The data makes a strong case. Team USA enters the 2028 Los Angeles Games with
four structural advantages stacking on top of each other.

**The baseline is already elite.** Since 1960, Team USA has averaged
`r round(us_baseline$mu)` gold medals per Summer Olympics more than any
other nation. The floor is high: even in tough Olympics, the US rarely falls
below 30 golds.

**Home soil matters.** Our analysis of every host nation since 1960 shows an
average gold medal increase of `r round(host_eff$mu, 0)`% when competing at
home. Crowd support, automatic qualification in every discipline, and years of
national investment leading up to the event all contribute. For Team USA, that
translates to roughly
`r round(us_baseline$mu * host_eff$mu / 100)` additional golds beyond the
baseline.

**The new sports are tailor-made for US athletes.** Baseball/softball, flag
football, lacrosse, and squash are all sports where American athletes have
dominated internationally. Historical data shows host nations win gold in
`r percent(prop_estimate, accuracy = 1)` of new disciplines they introduce.
With an estimated 10 gold-medal events across these sports, Team USA has a
meaningful cluster of additional wins available that no other country can match.

**The economic advantage is real.** With a GDP per capita of roughly
`r dollar(round(us_gdp_pc, -3))` well above the Olympic average of
`r dollar(round(avg_gdp_pc, -3))` the US investment in athlete development
reflects structural wealth that translates directly into medal performance.
Our regression confirms this relationship is statistically significant (p
< 0.001).

Putting it all together, the central estimate is
**`r round(gold_ci[2])` gold medals** for Team USA at LA 2028 (95% CI:
`r round(gold_ci[1])` to `r round(gold_ci[3])`). Adjusted for empirical model
error based on `r nrow(validation_hosts)` past host nations, the range is
**`r adjusted_lower` to `r adjusted_upper` golds**. That is a dominant
performance by any historical standard, and exactly the kind of outcome
sponsors want to be associated with.

---

::: {.callout-note title="AI Usage Statement" collapse="true"}

I used Claude (an AI assistant by Anthropic) to help structure and write the R
code for this project, including the web scraping pipeline in Task 1, the
Monte Carlo simulation in Task 5, and the bootstrap and GDP extra credit
sections. I reviewed all code outputs and interpreted the results throughout.
The written narrative, analysis framing, and conclusions are my own. No AI was
used to write or edit non-code text, in accordance with course policy.

:::
